---
title: "Mouse Cerebellum - Bipotent progenitors"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(Matrix)
  library(cluster)
  library(viridis)
  library(gridExtra)
  library(ArchR)
  library(RColorBrewer)
  library(ComplexHeatmap)
  library(SummarizedExperiment)
  library(pvclust)
})
```

```{r}
cpm_cutoff <- 5
pseudobulks <- read.table("~/Mouse_Cereb/014_Astroglia/Mouse_progenitors_by_timepoint_pseudobulk_cpm.txt", header = T, sep = "\t", stringsAsFactors = F)
```

A major question here is what is the relationship between the early Gsx1 positive cells (e10, e11, e12 - anterior VZ) and the later bipotent cells?

For this we need more targeted analyses.

1) Spearman's correlation using all active peaks

```{r}
pseudobulks.bip <- pseudobulks[, c("Progenitor_bipotent_e13", "Progenitor_bipotent_e15", colnames(pseudobulks)[grepl("e10|e11|e12", colnames(pseudobulks))])]

pseudobulks.bip.high <- pseudobulks.bip[apply(pseudobulks.bip, 1, max) > cpm_cutoff,]

bip_cors <- cor(pseudobulks.bip.high, method = "spearman")

bipotent_corrs <- bip_cors[c("Progenitor_bipotent_e13", "Progenitor_bipotent_e15"), grepl("e10|e11|e12", colnames(bip_cors))] %>%
  as.data.frame() %>%
  rownames_to_column("query") %>%
  gather(value = "corr", key = "subj", -query)

bipotent_corrs$timepoint <- sapply(bipotent_corrs$subj, function(x) str_split(x, "_", simplify=T)[length(str_split(x, "_", simplify=T))])

bipotent_corrs$type <- factor(sapply(1:nrow(bipotent_corrs), function(i) gsub(paste0("_", bipotent_corrs$timepoint[i]), "", bipotent_corrs$subj[i])), levels = c("Progenitor_anteriorVZ", "Progenitor", "Progenitor_isthmic", "Progenitor_VZ", "Progenitor_RL"))

p <- ggplot(bipotent_corrs, aes(x=type, y=corr, color=query)) +
  geom_point() +
  facet_wrap(~timepoint, scales = "free_y") +
  ylab("Spearman's rho") +
  xlab("") +
  theme_classic() +
  scale_color_manual(values = c("darkorange", "darkorchid4")) +
  theme(axis.text.x = element_text(angle=60, hjust = 1.1, vjust = 1.05), legend.position = "top")

p

pdf("~/Mouse_Cereb/Figures/006_astroglia/Bipotent_allPeaks_5cpm_corr_spearman.pdf", width = 6, height = 4.5, useDingbats = F); print(p); dev.off()
```

We see very strong temporal signals but overall the top hit is always the Gsx1 (anterior_VZ) progenitors (as well as uncommitted progenitors which might also contribute to the late bipotent ones) and isthmic progenitors which are at the border with the anterior VZ.

2) Marker peaks for bipotent progenitors vs other progenitors in e13-e15:

```{r}
## ArchR merged project
proj4 <- loadArchRProject("~/Mouse_Cereb/proj4_chromVar/")

cell_anno <- readRDS("~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cerebellum_atac_finalCellType_annotation.rds")

sum(cell_anno$cell == proj4$cellNames)

proj4$Cell_type_broad <- as.character(cell_anno$Cell_type_broad)
proj4$Cell_type <- as.character(cell_anno$Cell_type)

sort(table(cell_anno$Cell_type[cell_anno$Timepoint=="e13" & cell_anno$Cell_type_broad=="Astroglia"]), decreasing = T)
sort(table(cell_anno$Cell_type[cell_anno$Timepoint=="e15" & cell_anno$Cell_type_broad=="Astroglia"]), decreasing = T)

prog_cells <- c("Progenitor_bipotent", "Progenitor_RL", "Progenitor_isthmic", "Progenitor_gliogenic")

length(prog_cells)

e13_e15 <- subsetCells(proj4, cellNames = cell_anno$cell[cell_anno$Timepoint %in% c("e13", "e15")])

markersPeaks <- getMarkerFeatures(e13_e15, 
                                  groupBy = "Cell_type",
                                  useGroups = prog_cells,
                                  bgdGroups = prog_cells,
                                  useMatrix = "PeakMatrix",
                                  bias = c("TSSEnrichment", "log10(nFrags)")
                                  )

markerList <- getMarkers(markersPeaks, cutOff = "FDR <= 0.01 & Log2FC >= 1", returnGR = T)

heatmapPeaks <- markerHeatmap(
  seMarker = markersPeaks, 
  cutOff = "FDR <= 0.01 & Log2FC >= 1",
  transpose = TRUE
)

draw(heatmapPeaks, heatmap_legend_side = "bot", annotation_legend_side = "bot")
```

```{r}
pdf("~/Mouse_Cereb/Figures/006_astroglia/Bipotent_ee13_e15_ProgType_heatmap.pdf", width = 6, height = 4.5, useDingbats = F); draw(heatmapPeaks, heatmap_legend_side = "bot", annotation_legend_side = "bot");dev.off()
```


```{r}
bipotent_markers <- paste(seqnames(markerList[["Progenitor_bipotent"]]), start(markerList[["Progenitor_bipotent"]]), end(markerList[["Progenitor_bipotent"]]), sep = "_")

length(bipotent_markers)

sum(bipotent_markers %in% row.names(pseudobulks))

bipotent_markers <- bipotent_markers[bipotent_markers %in% row.names(pseudobulks)]
```

Fraction of reads in marker peaks

```{r}
bip_marker_f <- data.frame(
  subj=colnames(pseudobulks)[grepl("e10|e11|e12", colnames(pseudobulks))],
  f_CPM=colSums(pseudobulks[bipotent_markers, colnames(pseudobulks)[grepl("e10|e11|e12", colnames(pseudobulks))]])/colSums(pseudobulks[, colnames(pseudobulks)[grepl("e10|e11|e12", colnames(pseudobulks))]]), stringsAsFactors = F)

bip_marker_f$timepoint <- sapply(bip_marker_f$subj, function(x) str_split(x, "_", simplify=T)[length(str_split(x, "_", simplify=T))])

bip_marker_f$type <- factor(sapply(1:nrow(bip_marker_f), function(i) gsub(paste0("_", bip_marker_f$timepoint[i]), "", bip_marker_f$subj[i])), levels = c("Progenitor_anteriorVZ", "Progenitor_bipotent", "Progenitor", "Progenitor_isthmic", "Progenitor_VZ", "Progenitor_RL"))

p <- ggplot(bip_marker_f, aes(x=type, y=f_CPM, fill=timepoint)) +
  geom_col() +
  facet_wrap(~timepoint, scales = "free_y") +
  theme_classic() +
  xlab("") +
  ylab("Fraction of counts in bipotent marker peaks") +
  #scale_fill_manual(values = c("darkred","darkorange", "olivedrab")) +
  scale_fill_manual(values = brewer.pal(11, "Spectral")[1:3], guide=F) +
  theme(axis.text.x = element_text(angle=60, hjust = 1.1, vjust = 1.05), legend.position = "top")

p

pdf("~/Mouse_Cereb/Figures/006_astroglia/Bipotent_markerPeaks_e13-e15_fractionOfCounts_inEarlyProg.pdf", width = 6, height = 4.5, useDingbats = F); print(p); dev.off()
```

Based on these analyses, we clearly see that the Gsx1 positive progenitors from early stages (anterior VZ) are the best match to later bipotent progenitors, both in terms of overall correlations using all peaks (above a certain activity cutoff), but also when focusing on peaks that separate the bipotent progenitors from other stage-matched progenitor subtypes.

```{r}
if (dir.exists("~/Mouse_Cereb/014_Astroglia/Bipotent")==F) {
  dir.create("~/Mouse_Cereb/014_Astroglia/Bipotent")
}

saveRDS(markersPeaks, "~/Mouse_Cereb/014_Astroglia/Bipotent/markers_Peaks_prog_e13-e15.rds")
saveRDS(markerList, "~/Mouse_Cereb/014_Astroglia/Bipotent/markers_Peaks_prog_e13-e15_FDR0.01_LogFC_1.rds")
```

Let's see if we can also generate these statistics on a per cell basis:

```{r}
atac.seurat <- readRDS("~/Mouse_Cereb/014_Astroglia/Mouse_progenitors_Seurat.rds")

bip_query <- sapply(bipotent_markers, function(x) gsub("_", "-" , x))

f_bip <- Matrix::colSums(atac.seurat@assays$Peaks@counts[bip_query,])/Matrix::colSums(atac.seurat@assays$Peaks@counts)

f_bip_stats <- data.frame(timepoint=atac.seurat$timepoint,
                          Cell_type=atac.seurat$Cell_type,
                          f=f_bip,
                          stringsAsFactors = F) %>%
  filter(timepoint %in% c("e10", "e11", "e12"), Cell_type %in% c("Progenitor_anteriorVZ", "Progenitor", "Progenitor_isthmic", "Progenitor_VZ", "Progenitor_RL"))

f_bip_stats$Cell_type <- factor(f_bip_stats$Cell_type, levels = c("Progenitor_anteriorVZ", "Progenitor", "Progenitor_isthmic", "Progenitor_VZ", "Progenitor_RL"))

p <- ggplot(f_bip_stats, aes(x=Cell_type, y=f, fill=Cell_type)) +
  geom_boxplot(notch = T) +
  facet_wrap(~timepoint, scales = "free_y") +
  theme_classic() +
  xlab("") +
  ylab("Fraction of counts in bipotent-specific CREs") +
  #scale_fill_manual(values = c("darkred","darkorange", "olivedrab")) +
  scale_fill_manual(values = c("#795ad6", "#d9b99a", "#da6162", "#A1BBE1", "#6bb342"), guide=F) +
  theme(axis.text.x = element_text(angle=60, hjust = 1.1, vjust = 1.05), legend.position = "top")

p

pdf("~/Mouse_Cereb/Figures/006_astroglia/Bipotent_markerPeaks_e13-e15_fractionOfCounts_inEarlyProg_perCell.pdf", width = 6, height = 4, useDingbats = F); print(p); dev.off()

wilcox.test(f_bip_stats$f[f_bip_stats$Cell_type=="Progenitor_anteriorVZ" & f_bip_stats$timepoint=="e10"],
            f_bip_stats$f[f_bip_stats$Cell_type=="Progenitor" & f_bip_stats$timepoint=="e10"])

wilcox.test(f_bip_stats$f[f_bip_stats$Cell_type=="Progenitor_anteriorVZ" & f_bip_stats$timepoint=="e11"],
            f_bip_stats$f[f_bip_stats$Cell_type=="Progenitor" & f_bip_stats$timepoint=="e11"])

wilcox.test(f_bip_stats$f[f_bip_stats$Cell_type=="Progenitor_anteriorVZ" & f_bip_stats$timepoint=="e12"],
            f_bip_stats$f[f_bip_stats$Cell_type=="Progenitor" & f_bip_stats$timepoint=="e12"])

```

```{r}
sessionInfo()
```

