---
title: "Mouse Cerebellum 012 - Validate with RNA-seq"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
library(tidyverse)
library(RColorBrewer)
library(Seurat)
```

First, we want to see whether our cell type annotation is validated based on the integration with the data from Vladoiu et al. Nature 2019.

```{r}
atac_anno <- readRDS("~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cerebellum_atac_finalCellType_annotation.rds")

atac_anno$Cell_type <- as.character(atac_anno$Cell_type)

atac_anno$Cell_type[atac_anno$Cell_type=="Progenitor_bipotent" & atac_anno$Timepoint %in% c("e10", "e11", "e12")] <- "Progenitor_anteriorVZ"

atac_anno$Cell_type <- factor(atac_anno$Cell_type, levels = unique(atac_anno$Cell_type))

vladoiu_int <- list.files("~/Mouse_Cereb/011_integrate_RNAseq/Vladoiu_2019/Vladoiu_integration/", pattern = "summary.txt")
vladoiu_anno <- Reduce(bind_rows,lapply(vladoiu_int, function(x) {
  anno <- read.table(paste0("~/Mouse_Cereb/011_integrate_RNAseq/Vladoiu_2019/Vladoiu_integration/", x), header = T, sep = "\t", stringsAsFactors = F)
  anno$Vladoiu_timepoint <- str_split(x, "_", simplify = T)[7]
  anno$timepoint <- str_split(x, "_", simplify = T)[3]
  return(anno)
}))
```

Comparing annotations (we will only keep cells with annotation in Vladoiu, i.e. e10, e12, e14, e16, e18, P0, P7) and only cells with a prediction score >=0.9

```{r}
atac_anno_vladoiu <- dplyr::select(atac_anno, cellNames_short, Cell_type, Cell_type_broad, Cell_type_color, Cell_type_broad_color, Timepoint, Lib, Sex) %>%
  dplyr::rename(cell=cellNames_short) %>%
  inner_join(dplyr::select(vladoiu_anno,cell, Vladoiu_cell_type, Vladoiu_cell_type_predScore, Vladoiu_timepoint)) %>%
  filter(is.na(Vladoiu_cell_type)==F,
         Vladoiu_cell_type_predScore >= 0.9)

nrow(atac_anno_vladoiu)
```

Now let's look at the overlap between the annotations.

We will calculate a Jaccard index between all possible overlaps. For plotting, we only keep sets showing at least JI >= 0.15 with another set.

```{r, fig.width=8, fig.height=6}
atac_vlad_original_ji <- Reduce(bind_rows,lapply(unique(atac_anno_vladoiu$Cell_type), function(atac_c) {
  df <- Reduce(bind_rows,lapply(unique(atac_anno_vladoiu$Vladoiu_cell_type), function(vlad_c) {
    ji <- sum(atac_anno_vladoiu$Cell_type == atac_c & atac_anno_vladoiu$Vladoiu_cell_type ==vlad_c)/sum(atac_anno_vladoiu$Cell_type == atac_c | atac_anno_vladoiu$Vladoiu_cell_type ==vlad_c)
    df <- data.frame(Vladoiu_cell_type=vlad_c, JI=ji, stringsAsFactors = F)
    return(df)
  }))
  df$atac_cell_type <- atac_c
  return(df)
}))

atac_vlad_original_ji_mat <- spread(atac_vlad_original_ji, key = atac_cell_type, value = JI)
row.names(atac_vlad_original_ji_mat) <- atac_vlad_original_ji_mat$Vladoiu_cell_type
atac_vlad_original_ji_mat <- atac_vlad_original_ji_mat[,2:ncol(atac_vlad_original_ji_mat)]

summary(apply(atac_vlad_original_ji_mat, 1, max))
summary(apply(atac_vlad_original_ji_mat, 2, max))

pheatmap::pheatmap(atac_vlad_original_ji_mat, clustering_method = "ward.D2", color = colorRampPalette((brewer.pal(n = 9, name ="Blues")))(100))

i <- apply(atac_vlad_original_ji_mat, 1, max) >= 0.15
j <- apply(atac_vlad_original_ji_mat, 2, max) >= 0.15

pheatmap::pheatmap(atac_vlad_original_ji_mat[i,j], clustering_method = "ward.D2", color = colorRampPalette((brewer.pal(n = 9, name ="Blues")))(100))


```

```{r, fig.width=8, fig.height=6}
atac_vlad_original_ji_mat<-atac_vlad_original_ji_mat[i,j]

row.names(atac_vlad_original_ji_mat)
colnames(atac_vlad_original_ji_mat)

atac_vlad_original_ji_mat <- atac_vlad_original_ji_mat[
  c("Neural stem cells", "VZ progenitors", "Brainstem progenitors",
    "Gliogenic progenitors-1", "Gliogenic progenitors-2","Astrocyte/Bergmann glia precursors",
    "Upper rhombic lip progenitors",
    "Excitatory cerebellar nuclei neurons",
    "Post mitotic NTZ neurons",
    "Postnatal excitatory cerebellar nuclei neurons",
    "Differentiating Purkinje cells", "Purkinje cells",
    "GABA interneurons",
    "Embryonic and postnatal GCPs-1","Unipolar brush cell and GCP progenitor","Unipolar brush cell precursors",
    "Postnatal GCPs-2",
    "Unipolar brush cells",
    "Granule cells",
    "Oligodendrocyte precursor cells",
    "Microglia",
    "Meninges", "Pericytes",
    "Red blood cells"
  ),
  c("Progenitor", "Progenitor_anteriorVZ",   "Progenitor_VZ",
    "Progenitor_bipotent", "Progenitor_gliogenic", "astroblast_WM", "astroblast_GCL",
    "Parabrachial_isthmic_diff",
    "RL_Nb","Isthmic_nuclei", "Glut_DN_V_mature", "Glut_DN_P_mature", "GABA_DN_mature",
    "PC_diff", "PC_mature",
    "Interneuron_early",
    "GCP_prenatal", "GC_diff_prenatal", "UBC_diff",
    "GCP_postnatal", "GC_diff_P4P7", "GC_mature_P4P7", 
    "OPC",
    "Microglia",
    "Mural/Endothelial",
    "Erythroid"
  )
  ]

pheatmap::pheatmap(atac_vlad_original_ji_mat, cluster_rows = F, cluster_cols = F, color = colorRampPalette((brewer.pal(n = 9, name ="Blues")))(100), lwd=0.8)

pheatmap::pheatmap(atac_vlad_original_ji_mat, cluster_rows = F, cluster_cols = F, color = colorRampPalette((brewer.pal(n = 9, name ="Blues")))(100), lwd=0.8, filename = "~/Mouse_Cereb/Figures/002_finalNucSet/Mouse_atac_CellTypeAnno_withVladoiu_JI.pdf", width = 8, height = 6, cellheight = 10, cellwidth = 10)
```

Overall, the two annotations agree well. 
Exceptions: 
1) Our dif. PCs match brainstem progenitors, which should not be in our dataset given that we only dissected cerebella primordia (vs Vladoiu dissecting entire hindbrain).

2) GABA DN (not annotated in Vladoiu) also match the excitatory neurons. Indeed a small population in the Vlaodiu dataset shows Sox14/Gad2 activity.

```{r}
gene.symbols <- read_tsv("~/Data/scATAC_pipeline_v3/resources/mouse_coding_ens94_gene_symbols.txt")

table(vladoiu_anno$Vladoiu_timepoint[vladoiu_anno$Vladoiu_cell_type=="Excitatory cerebellar nuclei neurons"])

vladoiu.E14 <- readRDS("~/Mouse_Cereb/011_integrate_RNAseq/Vladoiu_2019/Vladoiu_rna_seurat/Vladoiu_E14_rna_seurat.rds")

DimPlot(vladoiu.E14, cells.highlight = row.names(vladoiu.E14@meta.data)[vladoiu.E14$Vladoiu_cell_type=="Excitatory cerebellar nuclei neurons"])

FeaturePlot(vladoiu.E14, gene.symbols$ensembl[gene.symbols$symbol=="Slc17a6"])
FeaturePlot(vladoiu.E14, gene.symbols$ensembl[gene.symbols$symbol=="Meis2"])
FeaturePlot(vladoiu.E14, gene.symbols$ensembl[gene.symbols$symbol=="Gad2"])
FeaturePlot(vladoiu.E14, gene.symbols$ensembl[gene.symbols$symbol=="Sox14"])

vladoiu.E14.df <- data.frame(UMAP1=vladoiu.E14@reductions$umap@cell.embeddings[,1],
                            UMAP2=vladoiu.E14@reductions$umap@cell.embeddings[,2],
                            Cell_type=vladoiu.E14$Vladoiu_cell_type, 
                            Slc17a6=vladoiu.E14@assays$RNA@counts[gene.symbols$ensembl[gene.symbols$symbol=="Slc17a6"],],
                            Meis2=vladoiu.E14@assays$RNA@counts[gene.symbols$ensembl[gene.symbols$symbol=="Meis2"],],
                            Gad2=vladoiu.E14@assays$RNA@counts[gene.symbols$ensembl[gene.symbols$symbol=="Gad2"],],
                            Sox14=vladoiu.E14@assays$RNA@counts[gene.symbols$ensembl[gene.symbols$symbol=="Sox14"],],
                            stringsAsFactors = F)

vladoiu.E14.df$Slc17a6 <- vladoiu.E14.df$Slc17a6 /max(vladoiu.E14.df$Slc17a6)
vladoiu.E14.df$Meis2 <- vladoiu.E14.df$Meis2 /max(vladoiu.E14.df$Meis2)
vladoiu.E14.df$Gad2 <- vladoiu.E14.df$Gad2 /max(vladoiu.E14.df$Gad2)
vladoiu.E14.df$Sox14 <- vladoiu.E14.df$Sox14 /max(vladoiu.E14.df$Sox14)

p1 <- ggplot(filter(vladoiu.E14.df, is.na(Cell_type)==F), aes(x=UMAP1, y=UMAP2, color=as.factor(Cell_type=="Excitatory cerebellar nuclei neurons"))) +
  geom_point(size=0.2, alpha=0.5) +
  theme_classic() +
  geom_rect(aes(xmin=4, xmax=19, ymin=-9, ymax=6), alpha=0, color="darkviolet", lwd=0.2, lty="dashed") +
  scale_color_manual(values = c("gray70", "darkviolet"), guide=F)

p1 

pdf("~/Mouse_Cereb/Figures/002_finalNucSet/Vladoiu_ExcitatoryNeurons_E14.pdf", width = 4, height = 4, useDingbats = F); print(p1); dev.off()

p2 <- ggplot(filter(vladoiu.E14.df, Cell_type=="Excitatory cerebellar nuclei neurons",UMAP1 > 4, UMAP1 < 19, UMAP2 <6, UMAP2 > (-9)), aes(x=UMAP1, y=UMAP2)) +
  geom_point(aes(alpha=Sox14),size=0.2, color="firebrick1") +
  geom_point(aes(alpha=Meis2),size=0.2, color="dodgerblue") +
  geom_point(aes(alpha=Gad2),size=0.2, color="gold2") +
  annotate("text", label="italic('Sox14')", x=5, y=-5 , color="firebrick1", parse=T, hjust=0) +
  annotate("text", label="italic('Meis2')", x=5, y=-4 , color="dodgerblue", parse=T, hjust=0) +
  annotate("text", label="italic('Gad2')", x=5, y=-3 , color="gold2", parse=T, hjust=0) +
  theme_classic() +
  theme(legend.position = "none")


p2

pdf("~/Mouse_Cereb/Figures/002_finalNucSet/Vladoiu_ExcitatoryNeurons_E14_Sox14_Meis2_Gad2.pdf", width = 3, height = 3, useDingbats = F); print(p2); dev.off()
```


2) A lot of our GCs are annotated as UBCs by Vladoiu. However, reanalysis of the scRNA-seq data revealed intracluster heterogeneity, where only a small fraction of the putative UBCs seems positive for UBC markers (Lmx1a, Eomes), whereas most cells are likely GCPs.

```{r}
table(vladoiu_anno$Vladoiu_timepoint[vladoiu_anno$Vladoiu_cell_type=="Unipolar brush cells"])

vladoiu.P0 <- readRDS("~/Mouse_Cereb/011_integrate_RNAseq/Vladoiu_2019/Vladoiu_rna_seurat/Vladoiu_P0_rna_seurat.rds")

DimPlot(vladoiu.P0, cells.highlight = row.names(vladoiu.P0@meta.data)[vladoiu.P0$Vladoiu_cell_type=="Unipolar brush cells"])

FeaturePlot(vladoiu.P0, gene.symbols$ensembl[gene.symbols$symbol=="Lmx1a"])
FeaturePlot(vladoiu.P0, gene.symbols$ensembl[gene.symbols$symbol=="Kcnd2"])

vladoiu.P0.df <- data.frame(UMAP1=vladoiu.P0@reductions$umap@cell.embeddings[,1],
                            UMAP2=vladoiu.P0@reductions$umap@cell.embeddings[,2],
                            Cell_type=vladoiu.P0$Vladoiu_cell_type, 
                            Lmx1a=vladoiu.P0@assays$RNA@counts[gene.symbols$ensembl[gene.symbols$symbol=="Lmx1a"],],
                            Kcnd2=vladoiu.P0@assays$RNA@counts[gene.symbols$ensembl[gene.symbols$symbol=="Kcnd2"],], stringsAsFactors = F)

vladoiu.P0.df$Lmx1a <- vladoiu.P0.df$Lmx1a /max(vladoiu.P0.df$Lmx1a)
vladoiu.P0.df$Kcnd2 <- vladoiu.P0.df$Kcnd2 /max(vladoiu.P0.df$Kcnd2)

p1 <- ggplot(filter(vladoiu.P0.df, is.na(Cell_type)==F), aes(x=UMAP1, y=UMAP2, color=as.factor(Cell_type=="Unipolar brush cells"))) +
  geom_point(size=0.2, alpha=0.5) +
  theme_classic() +
  geom_rect(aes(xmin=-3, xmax=7, ymin=-5, ymax=5), alpha=0, color="darkviolet", lwd=0.2, lty="dashed") +
  scale_color_manual(values = c("gray70", "darkviolet"), guide=F)

p1 

pdf("~/Mouse_Cereb/Figures/002_finalNucSet/Vladoiu_UnipolarBrushCells_P0.pdf", width = 4, height = 4, useDingbats = F); print(p1); dev.off()

p2 <- ggplot(filter(vladoiu.P0.df, Cell_type=="Unipolar brush cells",UMAP1 > (-3), UMAP1 < 7, UMAP2 <5, UMAP2 > (-5)), aes(x=UMAP1, y=UMAP2)) +
  geom_point(aes(alpha=Lmx1a),size=0.2, color="firebrick1") +
  geom_point(aes(alpha=Kcnd2),size=0.2, color="dodgerblue") +
  annotate("text", label="italic('Lmx1a')", x=-2.5, y=-2 , color="firebrick1", parse=T, hjust=0) +
  annotate("text", label="italic('Kcnd2')", x=-2.5, y=-3 , color="dodgerblue", parse=T, hjust=0) +
  theme_classic() +
  theme(legend.position = "none")


p2

pdf("~/Mouse_Cereb/Figures/002_finalNucSet/Vladoiu_UnipolarBrushCells_P0_Lmx1a_red_Kcn2_blue.pdf", width = 3, height = 3, useDingbats = F); print(p2); dev.off()
```

Thus, we can explain very well the inconsistencies in the two annotations.


Finally, let's look into the imputed RNA-seq data and how they compare with our gene activity estimates.

```{r}
vladoiu_cor_f <- list.files("~/Mouse_Cereb/011_integrate_RNAseq/Vladoiu_2019/Vladoiu_integration/", pattern = "acrossClust.txt")
vladoiu_cor <- Reduce(bind_rows,lapply(vladoiu_cor_f, function(x) {
  anno <- read.table(paste0("~/Mouse_Cereb/011_integrate_RNAseq/Vladoiu_2019/Vladoiu_integration/", x), header = T, sep = "\t", stringsAsFactors = F)
  anno$Vladoiu_timepoint <- str_split(x, "_", simplify = T)[7]
  anno$timepoint <- str_split(x, "_", simplify = T)[3]
  return(anno)
}))

vladoiu_cor$timepoint <- factor(vladoiu_cor$timepoint, levels = c("e10", "e12", "e13", "e15", "e17", "P0", "P4", "P7", "P14"))

p <- ggplot(vladoiu_cor, aes(GS_impRNA_corr, color=timepoint)) +
  geom_density() +
  scale_color_brewer(palette = "Spectral", name="Stage") +
  geom_vline(xintercept = median(na.omit(vladoiu_cor$GS_impRNA_corr)), lty="dashed", color="gray50") +
  xlab("Correlation between gene score and imputed expression (Pearson's r)") +
  theme_classic()

median(na.omit(vladoiu_cor$GS_impRNA_corr))

p

pdf("~/Mouse_Cereb/Figures/002_finalNucSet/Mouse_Cerebellum_GeneScore_vs_ImputedExpression_HVG_acrossClusters_Vladoiu.pdf", width = 5, height = 4, useDingbats = F); print(p); dev.off()
```

Prediction scores

```{r}
vladoiu_pred_f <- list.files("~/Mouse_Cereb/011_integrate_RNAseq/Vladoiu_2019/Vladoiu_integration/", pattern = "summary.txt")
vladoiu_pred <- unlist(sapply(vladoiu_pred_f, function(x) {
  anno <- read.table(paste0("~/Mouse_Cereb/011_integrate_RNAseq/Vladoiu_2019/Vladoiu_integration/", x), header = T, sep = "\t", stringsAsFactors = F)
  x <- as.numeric(anno$Vladoiu_cell_type_predScore)
  return(x)
}))

summary(vladoiu_pred)
```

```{r}
sessionInfo()
```

