---
title: "Mouse Cerebellum 002 - Final cell filtering"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ArchR)
  library(tidyverse)
  library(RColorBrewer)
})
```

We previously identified 1st and 2nd pass putative doublets.

Now we need to decide whether we will also remove clusters that were enriched for doublets/low quality cells.

This decision will be based on the marker genes that define these clusters:

```{r}
## Working directory. This is where the arrow files will be generated and stored.
setwd("~/Mouse_Cereb/")
proj <- loadArchRProject("proj1_init")
markersGS_clusters_1st_pass <- readRDS("002_doublets/Mou_Cer_1stPass_Cluster_markerGenes.rds")
markersGS <- getMarkers(markersGS_clusters_1st_pass)
put_doublets <- readLines("002_doublets/Mou_Cer_additional_put_doublets_Frag45K_DoubEnr4.txt")
spurious_cells <- readLines("002_doublets/Mou_Cer_cells_in_clustersEnriched_in_putDoublets.txt")
spurious_clust <- readLines("002_doublets/Mou_Cer_clustersEnriched_in_putDoublets.txt")
```

```{r, fig.width=8, fig.height=8}
plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Sample_short", embedding = "UMAP_qc", pal = rev(colorRampPalette(brewer.pal(11, "Spectral"))(22)), rastr = T, labelMeans = F)
plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Clusters_qc", embedding = "UMAP_qc", rastr = T)
```

No markers for C21! -> Remove!

Markers for C55

```{r}
p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name = markersGS[["C55"]]$name[1:30], 
    embedding = "UMAP_qc",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
  )

p
```

Remove -> doublets and low quality cells

Markers for C5

```{r}
p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name = markersGS[["C5"]]$name[1:30], 
    embedding = "UMAP_qc",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
  )

p
```

Remove -> Doublets between astroglia and GCs, interneurons!

Markers for C32

```{r}
p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name = markersGS[["C32"]]$name[1:30], 
    embedding = "UMAP_qc",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
  )

p
```

This is more tricky. Let's keep the cluster and only remove the high fragment/ doublet enriched cells.

Markers for C18

```{r}
p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name = markersGS[["C18"]]$name[1:30], 
    embedding = "UMAP_qc",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
  )

p
```

Clearly doublets, containing markers from different cell types

Markers for C54

```{r}
p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name = markersGS[["C54"]]$name[1:30], 
    embedding = "UMAP_qc",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
  )

p
```

These seem like genuine Purkinje cells. Removing the high fragment - doublet enriched cells should be enough!

Clusters to remove:

```{r}
clust_rm <- c("C21", "C55", "C5", "C18")
```

Now we can create a new project with the final set of filtered cells:

```{r}
subsetArchRProject(proj,
                   cells = proj$cellNames[!(proj$cellNames %in% c(proj$cellNames[proj$Clusters_qc %in% clust_rm], put_doublets))],
                   outputDirectory = "proj2_finalCells")

proj <- loadArchRProject("~/Mouse_Cereb/proj2_finalCells/")
```

Let's also update our barcode stats file:

```{r}
barcodes_init <- readRDS("001_barcode_qc/Mouse_Cerebellum_barcode_stats.rds")
barcodes_init$Spurious_clust <- barcodes_init$Clusters_qc %in% spurious_clust
barcodes_init$Spurious_cell_RM <- barcodes_init$Clusters_qc %in% clust_rm
barcodes_init$PassQC_all <- barcodes_init$barcode %in% proj$cellNames

saveRDS(barcodes_init, "001_barcode_qc/Mouse_Cerebellum_barcode_stats.rds")
```

```{r}
length(proj$cellNames)
```

Now we can generate some QC stats and plots:

TSS ridge plot

```{r}
p <- plotGroups(
    ArchRProj = proj, 
    groupBy = "Sample_short", 
    colorBy = "cellColData", 
    name = "TSSEnrichment",
    pal = rev(colorRampPalette(brewer.pal(11, "Spectral"))(22)),
    plotAs = "ridges"
   ) +
  ylab("Samples")

p

pdf("Figures/001_QC/TSSenrichment_byLib.pdf", width = 6, height = 6, useDingbats = F); print(p);dev.off()
```

Fragment size distribution

```{r}
fragment_sizes <- plotFragmentSizes(ArchRProj = proj, returnDF = T)

head(fragment_sizes)
dim(fragment_sizes)

fragment_sizes <- as.data.frame(fragment_sizes) %>%
  dplyr::rename(Sample=sampleName) %>%
  left_join(unique(dplyr::select(as.data.frame(getCellColData(proj)), Sample, Sample_short)))

p <-ggplot(fragment_sizes, aes(x=fragmentSize, y=fragmentPercent, color=as.factor(Sample_short))) +
  geom_line() +
  theme_classic() +
  xlab("Fragment size") +
  ylab("Density") +
  theme(legend.position = "bottom")+
  scale_color_manual(values = colorRampPalette(brewer.pal(11, "Spectral"))(22), name="Sample")

p

pdf("Figures/001_QC/FragLength_distribution_byLib.pdf", width = 6, height = 6, useDingbats = F); print(p);dev.off()

```

TSS enrichment

```{r}
tssEnr <- plotTSSEnrichment(ArchRProj = proj, returnDF = T)

head(tssEnr)
dim(tssEnr)

tssEnr <- as.data.frame(tssEnr) %>%
  dplyr::rename(Sample=sampleName) %>%
  left_join(unique(dplyr::select(as.data.frame(getCellColData(proj)), Sample, Sample_short)))

p <-ggplot(tssEnr, aes(x=x, y=smoothValue, color=as.factor(Sample_short))) +
  geom_line() +
  theme_classic() +
  xlab("Distance from TSS (bp)") +
  ylab("Normalised signal") +
  theme(legend.position = "bottom")+
  scale_color_manual(values = colorRampPalette(brewer.pal(11, "Spectral"))(22), name="Sample")

p

pdf("Figures/001_QC/TSS_signal_distribution_byLib.pdf", width = 6, height = 6, useDingbats = F); print(p);dev.off()
```

Number of cells per library

```{r}
p <- ggplot(filter(barcodes_init, PassQC_all==T), aes(Timepoint, fill=Sex)) +
  geom_bar() +
  coord_flip() +
  scale_fill_manual(values=c("darkorange", "deepskyblue3", "mediumorchid")) +
  xlab("Developmental stage") +
  ylab("Profiled nuclei") +
  theme_classic()

p

pdf("Figures/001_QC/Nuclei_final_by_Timepoint_and_Lib.pdf", width = 4, height = 4, useDingbats = F); print(p);dev.off()
```

Now we can rerun our iterative LSI in the final cell set:

```{r}
proj <- addIterativeLSI(ArchRProj = proj,
                        useMatrix = "TileMatrix",
                        name = "IterativeLSI_finalNucSet",
                        iterations=5,
                        clusterParams = list(
                          resolution = c(0.1, 0.2, 0.4, 0.8), 
                          sampleCells = 20000, 
                          n.start = 10
                        ), 
                        varFeatures = 100000,
                        dimsToUse = 1:100,
                        totalFeatures = 500000,
                        seed = 1,
                        LSIMethod = 1,
                        scaleDims = FALSE,
                        corCutOff = 0.75,
                        excludeChr = c("chrX", "chrY", "chrMT"),
                        binarize = T,
                        force = T)
```

Adding Clusters. Here we aim for broader cell type characterisation. A resolution of 1.5 seems to work best (we will later have to subcluster within cell types and developmental states).

```{r}
proj <- addClusters(input = proj,
                    name = "Clusters_finalNucSet_res1.5",
                    reducedDims = "IterativeLSI_finalNucSet",
                    method = "Seurat",
                    force = T,
                    resolution=1.5,
                    corCutOff = 0.75,
                    scaleDims = FALSE,
                    seed = 1)
```

We can also run a UMAP:

```{r}
proj <- addUMAP(ArchRProj = proj,
                name = "UMAP_finalNucSet",
                reducedDims = "IterativeLSI_finalNucSet",
                minDist = 0.25,
                metric = "cosine",
                nNeighbors = 30,
                force = T,
                seed = 1,
                scaleDims = F,
                corCutOff = 0.75)
```

```{r, fig.width=8, fig.height=8}
p1 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Sample_short", embedding = "UMAP_finalNucSet", pal = colorRampPalette(brewer.pal(11, "Spectral"))(22), rastr = T, labelMeans = F)
p2 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Clusters_finalNucSet_res1.5", embedding = "UMAP_finalNucSet", rastr = T)

ggAlignPlots(p1, p2, type = "h")
```

```{r}
if (dir.exists("Figures/002_finalNucSet")==F){
  dir.create("Figures/002_finalNucSet")
}

pdf("Figures/002_finalNucSet/Mouse_atac_LSI_pass2_tiles_bySample.pdf", width = 8, height = 8, useDingbats = F); print(p1);dev.off()

pdf("Figures/002_finalNucSet/Mouse_atac_LSI_pass2_tiles_byCluster_res1.5.pdf", width = 8, height = 8, useDingbats = F); print(p2);dev.off()
```

We should save the LSI, UMAP and Cluster information:

```{r}
if (dir.exists("003_finalNucSet")==F){
  dir.create("003_finalNucSet")
}

## UMAP
saveRDS(getEmbedding(proj, embedding = "UMAP_finalNucSet", returnDF = F), "003_finalNucSet/Mouse_Cereb_finalNucSet_UMAP_merged_tiles.rds")

## LSI
saveRDS(getReducedDims(proj, reducedDims = "IterativeLSI_finalNucSet", returnMatrix = F), "003_finalNucSet/Mouse_Cereb_finalNucSet_LSI_merged_tiles.rds")

## Clusters and other cell data
clust <- getCellColData(proj) %>%
  as.data.frame() %>%
  rownames_to_column("cellName")

saveRDS(clust, "003_finalNucSet/Mouse_Cereb_finalNucSet_Clusters_merged_tiles_res1.5_and_barcodeStats.rds")
```

Now, we want to assign cell types to our clusters. For that we will need to identify marker genes.

```{r}
markersGS_clusters_broad_cell_types <-getMarkerFeatures(proj,
                  groupBy = "Clusters_finalNucSet_res1.5",
                  useMatrix = "GeneScoreMatrix", 
                  threads = 40,
                  scaleTo = 10^4)

saveRDS(markersGS_clusters_broad_cell_types, "003_finalNucSet/Mouse_Cereb_finalNucSet_Clusters_merged_tiles_res1.5_markerGenes.rds")
```

We will have to examine these manually in the next section.

Let's save the project for now:

```{r}
saveArchRProject(proj)
```

```{r}
sessionInfo()
```

