---
title: "Mouse Cerebellum 003 - Broad cell type annotation"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ArchR)
  library(tidyverse)
  library(RColorBrewer)
  library(viridis)
  library(ComplexHeatmap)
})
```

Now it's time to match our clusters to cell types based on their marker genes.

We have already done this several times and already have a good feeling about which cell types are marked by which genes.
Let's focus on the most prominent marker genes:

```{r}
setwd("~/Mouse_Cereb/")
proj <- loadArchRProject("proj2_finalCells")
markersGS_clusters_broad_cell_types <- readRDS("003_finalNucSet/Mouse_Cereb_finalNucSet_Clusters_merged_tiles_res1.5_markerGenes.rds")
markersGS_cellAnno <- getMarkers(markersGS_clusters_broad_cell_types, cutOff = "FDR < 0.01 & Log2FC > 1")
```

```{r, fig.width=8, fig.height=8}
interesting_genes <- c("Atoh1", "Sox2", "Foxp2", "Cdon", "Gad2", "Sox14", "Tlx3", "Lmx1a", "Eomes", "Pax2", "Top2a", "Ptf1a", "Kirrel2", "Rora", "Rorb", "Slc17a6", "Meis2", "Aqp4", "Olig2", "Cbln3", "Fgf3", "Grin2b", "Neurod1", "Nhlh1", "Gdf10", "Fabp7", "Map3k1", "Gli2", "Esrrg", "Sorcs3", "Cdh22", "Etv1", "Junb", "Mef2a", "Foxp1", "Cdh9", "March11", "Pvalb", "Lhx2", "Pax5", "Syndig1l","Gabra6", "Slc1a3", "Inpp5d")

markerGenes <- c("Hbb-y", "Hbb-bh1", "Slc4a1", "Inpp5d", "Cx3cr1", "Angptl8", "Dlc1", "Col4a2", "Atoh1", "Mdga1", "Btbd11", "Map3k1", "Gli2", "Mgat5", "Lmx1a", "Dclk3", "Etv1", "Cbln3", "Gabra6", "Pax2", "Lhx5", "Cacna2d3", "March11", "Pvalb", "Grin2b", "Reln", "Nfib", "Neurod1", "Nhlh1", "Grik2", "Shisa9", "Foxp2", "Skor2", "Dab1", "Car8", "Fgf3", "Foxp1", "Cdh9", "Sox14", "Zfhx3", "Zfhx4", "Pax5", "Grm5", "Slc17a6", "Ebf2", "Lmx1b", "Tlx3", "Isl1", "Isl2", "Lef1", "Meis2", "Lhx9",  "Nhlh2", "Neurog2", "Notch1", "Sox1", "Sox2", "Cyp26b1","Fabp7","Slc1a3", "Kirrel2", "Ptf1a", "Pax3", "Gpc5", "Rorb", "Rora", "Mobp", "Sox10", "Olig2", "Lims2", "Gdf10", "Aqp4", "Etv4", "Etv5", "Aldoc")

p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name = union(interesting_genes, markerGenes), 
    embedding = "UMAP_finalNucSet",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)

p
```

Let's start exploring and labelling the clusters.

We can also consult the labels from our previous annotation:

```{r}
prev_cells <- read_tsv("~/Mouse_Cerebellum/003_process_timepoints/005_annotate_cellTypes_all/Mouse_merged_seurat_cellTypes_allAssays_summary.txt")

inspect_clust <- function(clust,
                          archrproj=proj,
                          slot="Clusters_finalNucSet_res1.5",
                          markers=markersGS_cellAnno,
                          umap="UMAP_finalNucSet") {
  ## Gettting top 20 markers, excluding LncRNAs and miRNAs with non-informative names
  m <- markers[[clust]]$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]", markers[[clust]]$name)==F][1:20]
  print("Top 20 markers")
  print(m)
  print("--------------------")
  ## Comparing to our previous annotation - cell type
  ## Index of cells belonging to the cluster
  i <- which(getCellColData(archrproj, slot,drop = T)==clust)
  print("Comparison to previous annotation")
  print(sort(table(prev_cells$cell_type[prev_cells$cell %in% archrproj$cellNames_short[i]]), decreasing = T))
  print("--------------------")
  ## Timepoint distribution
  print("Timepoint distribution")
  print(sort(table(archrproj$Timepoint[i]), decreasing = T))
  print("--------------------")
  ## Plotting top genes
  print("Plotting top marker genes")
  p <- plotEmbedding(
    ArchRProj = archrproj, 
    colorBy = "GeneScoreMatrix", 
    name =m, 
    embedding = umap,
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL,
  )
  print(p)
  print("--------------------")
}

inspect_clust("C1")
## Erythroid
```

```{r}
inspect_clust("C2")
## Microglia
```

```{r}
inspect_clust("C3")
## Vascular (mural/endothelia/meninges)
```

```{r}
inspect_clust("C4")
## Oligodendrocytes
```

```{r}
inspect_clust("C5")
## GC subset - but no Atoh1, Gli2 but showing lots of mitotic markers -> labelling them as apoptotic GCs
```

```{r}
inspect_clust("C6")
## GCP prenatal
```

```{r}
inspect_clust("C7")
## GCP postnatal
```

```{r}
inspect_clust("C8")
## Postmitotic GCs P4-P7 (migratory)
```

```{r}
inspect_clust("C9")
## mature GCs P4-P7
```

```{r}
inspect_clust("C10")
## GCPs and migrating P14
```

```{r}
inspect_clust("C11")
## Interneurons
```

```{r}
inspect_clust("C12")
## Postmitotic prenatal GCs (inner EGL)
```

```{r}
inspect_clust("C13")
## UBCs differentiating
```

```{r}
inspect_clust("C14")
## Interneurons late
```

```{r}
inspect_clust("C15")
## UBCs adult
```

```{r}
inspect_clust("C16")
## GC IGL P14
```

```{r}
inspect_clust("C17")
## GC IGL P63
```

```{r}
inspect_clust("C18")
## GC IGL P63
```

```{r}
inspect_clust("C19")
## e10 progenitor
```

```{r}
inspect_clust("C20")
## e10 progenitor
```

```{r}
inspect_clust("C21")
## e12 progenitor
```

```{r}
inspect_clust("C22")
## e11 progenitor
```

```{r}
inspect_clust("C23")
plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name =c("Lmx1a", "Slc17a6", "Tlx3", "Isl1", "Lmx1b"), 
    embedding = "UMAP_finalNucSet",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)
# Seems like mixture of parabrachial & isthmic -> will need to recluster
```

```{r}
inspect_clust("C24")
## GABA DN e13-P0
```

```{r}
inspect_clust("C25")
## GABA DN e12
```

```{r}
inspect_clust("C26")
## PC e12
```

```{r}
inspect_clust("C27")
"Dab1" %in% markersGS_cellAnno[["C27"]]$name
"Ebf2" %in% markersGS_cellAnno[["C27"]]$name
"Fgf3" %in% markersGS_cellAnno[["C27"]]$name
"Cdh9" %in% markersGS_cellAnno[["C27"]]$name
## PC e15-e17 Ebf2
```

```{r}
inspect_clust("C28")
"Ebf2" %in% markersGS_cellAnno[["C28"]]$name
"Dab1" %in% markersGS_cellAnno[["C28"]]$name
"Fgf3" %in% markersGS_cellAnno[["C28"]]$name
"Cdh9" %in% markersGS_cellAnno[["C28"]]$name
## PC e15-e17 Dab1
```

```{r}
inspect_clust("C29")
"Ebf2" %in% markersGS_cellAnno[["C29"]]$name
"Dab1" %in% markersGS_cellAnno[["C29"]]$name
"Fgf3" %in% markersGS_cellAnno[["C29"]]$name
"Cdh9" %in% markersGS_cellAnno[["C29"]]$name
## PCs Ebf2
```

```{r}
inspect_clust("C30")
"Ebf2" %in% markersGS_cellAnno[["C30"]]$name
"Dab1" %in% markersGS_cellAnno[["C30"]]$name
"Fgf3" %in% markersGS_cellAnno[["C30"]]$name
"Cdh9" %in% markersGS_cellAnno[["C30"]]$name
## PCs Ebf2 Fgf3
```

```{r}
inspect_clust("C31")
## PC Nbs
```

```{r}
inspect_clust("C32")
## Mixture of PC (Skor2, Foxp2) and interneuron Nbs (Pax2, Neurod6), as well as isthmic mantle (Pax5, Pax8) -> Need to recluster
```

```{r}
inspect_clust("C33")
## PC Nbs
```

```{r}
inspect_clust("C34")

"Lhx9" %in% markersGS_cellAnno[["C34"]]$name
"Lmx1b" %in% markersGS_cellAnno[["C34"]]$name
"Tlx3" %in% markersGS_cellAnno[["C34"]]$name
"Lmo3" %in% markersGS_cellAnno[["C34"]]$name
"Epha4" %in% markersGS_cellAnno[["C34"]]$name
## Glut DNs (Meis2, Slc17a6, Neurod6 but no Lhx9 [isthmic Nuclei], Lmx1b [parabrachial], TLx3[MBO])
```

```{r}
inspect_clust("C35")

"Lhx9" %in% markersGS_cellAnno[["C35"]]$name
"Lmx1b" %in% markersGS_cellAnno[["C35"]]$name
"Tlx3" %in% markersGS_cellAnno[["C35"]]$name
"Lmo3" %in% markersGS_cellAnno[["C35"]]$name
"Epha4" %in% markersGS_cellAnno[["C35"]]$name
## Glut DN ventral and isthmic nuclei (Meis2, Slc17a6, Neurod6 and Lmo3 but also Lhx9, Evx1 [isthmic Nuclei]) -> can be resolved by subclustering!
```

```{r}
inspect_clust("C36")
## RL NBs: Glut DN and isthmic neuroblasts (e11)
```

```{r}
inspect_clust("C37")
## RL NBs: Glut DN and isthmic neuroblasts (e10)
```

```{r}
inspect_clust("C38")
## VZ NBs: GABA DN and MBO, parabrachial (e10)
```

```{r}
inspect_clust("C39")
## VZ NBs: GABA DN and MBO, parabrachial (e11)
```

```{r}
inspect_clust("C40")

"Gdf10" %in% markersGS_cellAnno[["C40"]]$name
"Olig2" %in% markersGS_cellAnno[["C40"]]$name
"Sox2" %in% markersGS_cellAnno[["C40"]]$name
## Bergmann astroglia precursors (+ Etv4, Etv5 although absent from marker gene list)
```

```{r}
inspect_clust("C41")

"Gdf10" %in% markersGS_cellAnno[["C41"]]$name
"Olig2" %in% markersGS_cellAnno[["C41"]]$name
"Sox2" %in% markersGS_cellAnno[["C41"]]$name
## Bergmann astroglia
```

```{r}
inspect_clust("C42")
## OPC
```

```{r}
inspect_clust("C43")
## Intermediate progenitors and interneuron neuroblasts
```

```{r}
inspect_clust("C44")
## Progenitors
```

```{r}
inspect_clust("C45")
## Progenitors e13 (Gsx1 + generating interneurons!)
```

```{r}
inspect_clust("C46")
## Progenitors e15
```

```{r}
inspect_clust("C47")
"Notch1" %in%  markersGS_cellAnno[["C47"]]$name
"Gdf10" %in%  markersGS_cellAnno[["C47"]]$name
## Bergmann precursors e17-P0
```

```{r}
markerGenes <- c("Hbb-y", "Hbb-bh1", "Slc4a1", "Inpp5d", "Cx3cr1", "Angptl8", "Dlc1", "Col4a2", "Atoh1", "Mdga1", "Btbd11", "Map3k1", "Gli2", "Mgat5", "Lmx1a", "Dclk3", "Etv1", "Cbln3", "Gabra6", "Pax2", "Lhx5", "Cacna2d3", "March11", "Pvalb", "Grin2b", "Reln", "Nfib", "Neurod1", "Nhlh1", "Grik2", "Shisa9", "Foxp2", "Skor2", "Dab1", "Car8", "Fgf3", "Foxp1", "Cdh9", "Sox14", "Zfhx3", "Zfhx4", "Pax5", "Grm5", "Slc17a6", "Ebf2", "Lmx1b", "Tlx3", "Isl1", "Isl2", "Lef1", "Meis2", "Lhx9",  "Nhlh2", "Neurog2", "Notch1", "Sox1", "Sox2", "Cyp26b1","Fabp7","Slc1a3", "Kirrel2", "Ptf1a", "Pax3", "Hes4","Gpc5", "Rorb", "Rora", "Mobp", "Sox10", "Olig2", "Lims2", "Gdf10", "Aqp4", "Etv4", "Etv5", "Aldoc")

length(markerGenes)
```

Making a heatmap:

```{r, fig.height=8, fig.width=8}
mark_stats <- as.matrix(assay(markersGS_clusters_broad_cell_types))
row.names(mark_stats) <-as.character(markersGS_clusters_broad_cell_types@elementMetadata$name)

top_markers <- unique(as.character(sapply(markersGS_cellAnno, function(x) as.character(x$name[1:5]))))

mark_stats.plot <- mark_stats[top_markers,]

## Cap to 2 /-1
mark_stats.plot[mark_stats.plot > 2] <- 2
mark_stats.plot[mark_stats.plot < (-1)] <- (-1)

marker_labels = rowAnnotation(foo = anno_mark(
  at = grep(paste(markerGenes, collapse="|"),row.names(mark_stats.plot)),
  labels = row.names(mark_stats.plot)[grepl(paste(markerGenes, collapse="|"), row.names(mark_stats.plot))]),
  annotation_legend_param = list(direction = "vertical",
                                 nrow = 1, labels_gp = gpar(fontsize = 0.1),
                                 padding = unit(10, "mm")))

draw(Heatmap(matrix = mark_stats.plot,
        cluster_rows = T,
        cluster_columns = T,
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_row_names = F,
        show_column_names = T,
        col=viridis(100,option = "B"),
        right_annotation = marker_labels,
        border = TRUE,
        heatmap_legend_param = list(direction = "vertical", title = "Z-score GS")), heatmap_legend_side = "right")
```

Cell type annotation:

```{r}
## Importing cluster - cell type annotation correspondence
cell_type_anno <- read_tsv("~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cer_arch_Clusters_finalNucSet_res1.5_C47_annotation.txt")

getCellColData(proj, "Clusters_finalNucSet_res1.5")

cell_type_anno.df <- getCellColData(proj, "Clusters_finalNucSet_res1.5", drop = F) %>%
  as.data.frame() %>%
  rownames_to_column("cellName") %>%
  left_join(cell_type_anno)

if(sum(cell_type_anno.df$cellName ==  getCellNames(proj))) {
  proj$Cell_type_broad_C47 <- as.character(cell_type_anno.df$Cell_type_broad)
  proj$Cell_type_C47 <- cell_type_anno.df$Cell_type
  proj$Subclustering_group_C47 <- cell_type_anno.df$Subclustering_group
}

table(proj$Cell_type_broad_C47)
```

Let's save the project for now:

```{r}
saveArchRProject(proj)
```

Now we can make subsets of this project based on the broad cell type annotation and repeat the clustering. 

For that, we will need to subset the original fragment files to save space:

```{r}
## This function will retrieve barcodes associated with specified [groups] in the [group_slot] for each sample and store them in a nested list object. The structure is: group > sample > barcodes
getNamedGroupList <- function(groups, group_slot, archr_proj=proj) {
  named_group_lists <- lapply(groups, function(g) {
    group_barcodes_all_samples <- archr_proj$cellNames[as.data.frame(getCellColData(archr_proj,group_slot))[,1] == g]
    mat <- do.call(rbind, lapply(group_barcodes_all_samples, function(x) {
      str_split(x, "#", simplify = T)
    }))
    group_barcodes_by_sample <- lapply(unique(mat[,1]), function(s) {
      mat[,2][mat[,1] == s]
    })
    names(group_barcodes_by_sample) <- unique(mat[,1])
    return(group_barcodes_by_sample)    
  })
  names(named_group_lists) <- groups
  return(named_group_lists)
}

## This function will go through the fragments of a single sample (condition/tissue/stage + replicate) and will extract the fragments that belong to each group.
## We are iterating over samples rather than groups because the most computationally demanding step is reading each fragment file.
## The object "named_group_list" needs to be constructed for a set of samples and groups with the function: getNamedGroupList().
## We follow a nested list structure with the format group > sample > barcodes. Barcodes must be a character vector with the barcodes that belong in this group for this sample, without any prefixes (i.e. as they appear in the fragments file).
splitFragments <- function(named_group_list, out_dir="~/Mouse_Cereb/005_reprocess_by_cellType_broad/", frag_dir="~/sds/sd17d003/scATAC/Cerebellum/Mouse/") {
  ## First, we need to identify the samples that are relevant for these groups.
  sample_names <- unique(unlist(sapply(named_group_list, names)))
  lapply(sample_names, function(sample_name) {
      ## Reading in the fragments
      frags <- fread(paste0(frag_dir, sample_name,"/cellranger/",sample_name,"/outs/fragments.tsv.gz"), header = F, nThread = 10)
      ## Splitting the fragments by cluster (will need to update the cluster slot for merged)
      lapply(names(named_group_list), function(g) {
        ## Filtering the fragments that belong to the particular cluster
        group_frag <- filter(frags, V4 %in% named_group_list[[g]][[sample_name]])
        ## Saving the fragments from the respective group and sample
        if (dir.exists(paste0(out_dir, g, "/fragments/"))==F){
         dir.create(paste0(out_dir, g, "/fragments/"), recursive = T)
        }
        write_tsv(group_frag, path = paste0(out_dir, g, "/fragments/", sample_name, "_fragments.tsv"), col_names = F)
        print(paste0("Exported ", nrow(group_frag), " fragments for group ", g, " from sample ", sample_name))
    })
  })
  return(NULL)
}
```

Now we specify the groups we are interested in subsetting and their specified slot in the cellColData container of the archr project.

```{r}
n_g_list <- getNamedGroupList(groups = unique(grep("Group", proj$Subclustering_group_C47, value = T)), group_slot = "Subclustering_group_C47", archr_proj = proj)

splitFragments(n_g_list)
```

Then we need to sort the fragment files:

```{bash}
for i in ~/Mouse_Cereb/005_reprocess_by_cellType_broad/*/fragments/*_fragments.tsv; do sort -V -k1,1 -k2,2n -k3,3n $i > ${i%.*}_sorted.tsv; done
```

Replacing the original files to save some space:

```{bash}
for i in ~/Mouse_Cereb/005_reprocess_by_cellType_broad/*/fragments/*_sorted*; do rename -f 's/_sorted//' $i; done
```

Gziping the files:

```{bash}
for i in ~/Mouse_Cereb/005_reprocess_by_cellType_broad/*/fragments/*_fragments.tsv; do bgzip $i; done
```

Finally, indexing the files with tabix:

```{bash}
for i in ~/Mouse_Cereb/005_reprocess_by_cellType_broad/*/fragments/*_fragments.tsv.gz; do tabix -s 1 -b 2 -e 3 $i; done
```

Now we are ready to reprocess each subset with ArchR (next sections).

```{r}
sessionInfo()
```

