---
title: "Mouse Cerebellum 004 - Reprocess subgroup"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ArchR)
  library(tidyverse)
  library(RColorBrewer)
  library(ComplexHeatmap)
  library(viridis)
})
```

Group to be reprocessed:

```{r}
subgroup <- "Group1_astroglia"

print(paste0("Reprocessing subgroup: ", subgroup))
```


Initialising ArchR

```{r}
addArchRThreads(threads = 20) 

## Working directory. This is where the arrow files will be generated and stored.
setwd(paste0("~/Mouse_Cereb/005_reprocess_by_cellType_broad/", subgroup))

## Genome and annotation
addArchRGenome("mm10")
```

```{r}
inputFiles <- list.files("fragments/", "_fragments.tsv.gz$")

inputFiles
names(inputFiles) <- sapply(inputFiles, function(x) gsub("_fragments.tsv.gz", "", x))
```

Now we can create our arrow files. We use the same QC metrics as before (TSS >=3, nFrags >= 5000) so all barcodes in the fragment files should pass QC:

```{r}
if (dir.exists("001_barcode_qc")==F){
  dir.create("001_barcode_qc")
}

ArrowFiles <- createArrowFiles(
  inputFiles = paste0("fragments/",inputFiles),
  sampleNames = names(inputFiles),
  filterTSS = 3, 
  filterFrags = 5000, 
  addTileMat = TRUE,
  addGeneScoreMat = TRUE,
  QCDir = "001_barcode_qc",
  promoterRegion = c(2000,100),
  removeFilteredCells = T,
  minFrags = 2500, 
  force = T
)
```

No need to calculate doublet scores or filter for doublets, these have already been removed from the merged dataset.

We can directly initialise the project.

```{r}
proj <- ArchRProject(
  ArrowFiles = ArrowFiles, 
  outputDirectory = paste0("proj_", subgroup),
  copyArrows = TRUE #This is recommened so that you maintain an unaltered copy for later usage.
)

## Also adding more metadata
proj$cellNames_short <- sapply(getCellNames(proj), function(x) {
  paste(str_extract(x, "sa[0-9]*"), str_split(x, pattern = fixed("#"), simplify = T)[2], sep="_")
})

proj$Sample_short <- sapply(proj$Sample, function(x) {
  x <- paste(str_split(x, pattern = fixed("_"), simplify = T)[c(4, 1, 5)], collapse="_")
  x <- gsub("P4", "P04", x)
  x <- gsub("P7", "P07", x)
})

proj$Timepoint <- sapply(proj$Sample_short, function(x) str_split(x, pattern = fixed("_"), simplify = T)[1])
proj$Lib <- sapply(proj$Sample_short, function(x) str_split(x, pattern = fixed("_"), simplify = T)[2])
proj$Sex <- sapply(proj$Sample_short, function(x) str_split(x, pattern = fixed("_"), simplify = T)[3])
```

Adding iterative LSI. Here we are analysing a more homogeneous set of cells compared to the entire dataset. Thus, we can reduce the number of dimensions and features to be used.

```{r}
proj <- addIterativeLSI(ArchRProj = proj,
                        useMatrix = "TileMatrix",
                        name = paste0("IterativeLSI_", subgroup),
                        iterations=5,
                        clusterParams = list(
                          resolution = c(0.1, 0.2, 0.4, 0.8), 
                          sampleCells = 10000, 
                          n.start = 10
                        ), 
                        varFeatures = 50000,
                        dimsToUse = 1:50,
                        totalFeatures = 200000,
                        seed = 1,
                        LSIMethod = 1,
                        scaleDims = FALSE,
                        corCutOff = 0.75,
                        excludeChr = c("chrX", "chrY", "chrMT"),
                        binarize = T,
                        force = T)
```

Adding clusters:

```{r}
proj <- addClusters(input = proj,
                    name = paste0("Clusters_within_", subgroup),
                    reducedDims = paste0("IterativeLSI_", subgroup),
                    method = "Seurat",
                    force = T,
                    resolution=2.5,
                    corCutOff = 0.75,
                    scaleDims = FALSE,
                    seed = 1)
```

And a UMAP. Here we are looking for substructure so we can reduce the number of neighbors:

```{r}
proj <- addUMAP(ArchRProj = proj,
                name = paste0("UMAP_within_", subgroup),
                reducedDims = paste0("IterativeLSI_", subgroup),
                minDist = 0.1,
                metric = "cosine",
                nNeighbors = 10,
                force = T,
                seed = 1,
                scaleDims = F,
                corCutOff = 0.75)
```

Plotting

```{r, fig.width=8, fig.height=8}
p1 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Sample_short", embedding = paste0("UMAP_within_", subgroup), pal = colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample))), rastr = T, labelMeans = F)
p2 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = paste0("Clusters_within_", subgroup), embedding = paste0("UMAP_within_", subgroup), rastr = T) + theme(legend.position = "none")

p1
p2
```

```{r, fig.width=5, fig.height=5}
interesting_genes <- c("Notch1","Sox2","Fabp7", "Cyp26b1","Id3", "Id4", "Hes5", "Nes", "Slc1a3", ## General astroglia
                       "Cdon","Atoh1", "Pax6", "Reln", "Barhl1", "Wnt1", ## RL
                        "Dll1","Kirrel2","Ptf1a",  "Neurog2","Tfap2b", "Msx3", "Mt3", ## VZ
                       "Gsx1", "Pax2", "Neurod6", ## VZ interneurons
                       "Olig2","Skor2", ## VZ PCPs
                       "Robo4", "Nhlh1", "Neurog2", #GABA Nbs
                       "Slc17a6", "Meis2", "Evx1", "Barhl1",#Glut Nbs
                       "Sox10", "Lims2", ## OPCs
                       "Gdf10", "Aqp4", "A2m", "Etv4", "Etv5", "Grm3","Tnc") ## Bergmann, astroglia)

interesting_genes <- interesting_genes[interesting_genes %in% proj@geneAnnotation$genes$symbol]

p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name = interesting_genes, 
    embedding = paste0("UMAP_within_", subgroup),
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)

p
```

Finding marker genes between the clusters

```{r}
markersGS_clusters <-getMarkerFeatures(proj,
                  groupBy = paste0("Clusters_within_", subgroup),
                  useGroups = names(which(table(getCellColData(proj, paste0("Clusters_within_", subgroup), drop = T)) > 50)),
                  useMatrix = "GeneScoreMatrix",
                  threads = 10,
                  scaleTo = 10^4)
markersGS <- getMarkers(markersGS_clusters, cutOff = "FDR < 0.01 & Log2FC > 0.5")
```

```{r, fig.height=12, fig.width=7, eval=FALSE}
mark_stats <- as.matrix(assay(markersGS_clusters))
row.names(mark_stats) <-as.character(markersGS_clusters@elementMetadata$name)

top_markers <- na.omit(unique(as.character(sapply(markersGS, function(x) as.character(x$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|^Olfr", x$name)==F][1:5])))))

mark_stats.plot <- mark_stats[top_markers,]

## Cap to 2 /-1
mark_stats.plot[mark_stats.plot > 2] <- 2
mark_stats.plot[mark_stats.plot < (-1)] <- (-1)

draw(Heatmap(matrix = mark_stats.plot,
        cluster_rows = T,
        cluster_columns = T,
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_row_names = T,
        show_column_names = T,
        col=viridis(100,option = "B"),
        border = TRUE,
        heatmap_legend_param = list(direction = "vertical", title = "Z-score GS")), heatmap_legend_side = "right")
```

Let's have a closer look into these clusters:

```{r}
prev_cells <- read_tsv("~/Mouse_Cerebellum/003_process_timepoints/005_annotate_cellTypes_all/Mouse_merged_seurat_cellTypes_allAssays_summary.txt")

inspect_clust <- function(clust,
                          archrproj=proj,
                          slot=paste0("Clusters_within_", subgroup),
                          markers=markersGS,
                          umap=paste0("UMAP_within_", subgroup)) {
  ## Gettting top 20 markers, excluding LncRNAs and miRNAs with non-informative names
  m <- markers[[clust]]$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|^Olfr", markers[[clust]]$name)==F][1:10]
  m <- na.omit(as.character(m))
  print("Top 10 markers")
  print(m)
  print("--------------------")
  ## Comparing to our previous annotation - cell type
  ## Index of cells belonging to the cluster
  i <- which(getCellColData(archrproj, slot,drop = T)==clust)
  print("Comparison to previous annotation")
  print(sort(table(prev_cells$cell_type[prev_cells$cell %in% archrproj$cellNames_short[i]]), decreasing = T))
  print("--------------------")
  ## Timepoint distribution
  print("Timepoint distribution")
  print(sort(table(archrproj$Timepoint[i]), decreasing = T))
  print("--------------------")
  ## Plotting top genes
  print("Plotting top marker genes")
  p <- plotEmbedding(
    ArchRProj = archrproj, 
    colorBy = "GeneScoreMatrix", 
    name =m, 
    embedding = umap,
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL,
  )
  print(p)
  print("--------------------")
}

lapply(sort(unique(proj$Clusters_within_Group1_astroglia)), inspect_clust)
```

We see that we have strong temporal effects. We should also consider calling markers by stage:

###e10

```{r}
clust_e10 <- names(which(table(proj$Clusters_within_Group1_astroglia[proj$Timepoint=="e10"]) > 50))

markersGS_clusters_e10 <-getMarkerFeatures(proj,
                  groupBy = paste0("Clusters_within_", subgroup),
                  useGroups = clust_e10,
                  bgdGroups = clust_e10,
                  useMatrix = "GeneScoreMatrix",
                  threads = 10,
                  scaleTo = 10^4)
markersGS_e10 <- getMarkers(markersGS_clusters_e10, cutOff = "FDR < 0.01 & Log2FC > 0.5")
```

```{r, fig.height=10, fig.width=5}
mark_stats_e10 <- as.matrix(assay(markersGS_clusters_e10))
row.names(mark_stats_e10) <-as.character(markersGS_clusters_e10@elementMetadata$name)

top_markers <- na.omit(unique(as.character(sapply(markersGS_e10, function(x) as.character(x$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|^Olfr", x$name)==F][1:10])))))

mark_stats_e10.plot <- mark_stats_e10[top_markers,]

## Cap to 2 /-1
mark_stats_e10.plot[mark_stats_e10.plot > 2] <- 2
mark_stats_e10.plot[mark_stats_e10.plot < (-1)] <- (-1)

print(draw(Heatmap(matrix = mark_stats_e10.plot,
        cluster_rows = T,
        cluster_columns = T,
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_row_names = T,
        show_column_names = T,
        col=viridis(100,option = "B"),
        border = TRUE,
        heatmap_legend_param = list(direction = "vertical", title = "Z-score GS")), heatmap_legend_side = "right"))
```

```{r, fig.width=5, fig.height=5}
lapply(clust_e10, function(c) {
  print(c)
  inspect_clust(c, markers = markersGS_e10)
})
```

###e11

```{r}
clust_e11 <- names(which(table(proj$Clusters_within_Group1_astroglia[proj$Timepoint=="e11"]) > 50))

markersGS_clusters_e11 <-getMarkerFeatures(proj,
                  groupBy = paste0("Clusters_within_", subgroup),
                  useGroups = clust_e11,
                  bgdGroups = clust_e11,
                  useMatrix = "GeneScoreMatrix",
                  threads = 10,
                  scaleTo = 10^4)
markersGS_e11 <- getMarkers(markersGS_clusters_e11, cutOff = "FDR < 0.01 & Log2FC > 0.5")
```

```{r, fig.height=10, fig.width=5}
mark_stats_e11 <- as.matrix(assay(markersGS_clusters_e11))
row.names(mark_stats_e11) <-as.character(markersGS_clusters_e11@elementMetadata$name)

top_markers <- na.omit(unique(as.character(sapply(markersGS_e11, function(x) as.character(x$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|^Olfr", x$name)==F][1:10])))))

mark_stats_e11.plot <- mark_stats_e11[top_markers,]

## Cap to 2 /-1
mark_stats_e11.plot[mark_stats_e11.plot > 2] <- 2
mark_stats_e11.plot[mark_stats_e11.plot < (-1)] <- (-1)

print(draw(Heatmap(matrix = mark_stats_e11.plot,
        cluster_rows = T,
        cluster_columns = T,
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_row_names = T,
        show_column_names = T,
        col=viridis(100,option = "B"),
        border = TRUE,
        heatmap_legend_param = list(direction = "vertical", title = "Z-score GS")), heatmap_legend_side = "right"))
```

```{r, fig.width=5, fig.height=5}
lapply(clust_e11, function(c) {
  print(c)
  inspect_clust(c, markers = markersGS_e11)
})
```

###e12

```{r}
clust_e12 <- names(which(table(proj$Clusters_within_Group1_astroglia[proj$Timepoint=="e12"]) > 50))

markersGS_clusters_e12 <-getMarkerFeatures(proj,
                  groupBy = paste0("Clusters_within_", subgroup),
                  useGroups = clust_e12,
                  bgdGroups = clust_e12,
                  useMatrix = "GeneScoreMatrix",
                  threads = 10,
                  scaleTo = 10^4)
markersGS_e12 <- getMarkers(markersGS_clusters_e12, cutOff = "FDR < 0.01 & Log2FC > 0.5")
```

```{r, fig.height=10, fig.width=5}
mark_stats_e12 <- as.matrix(assay(markersGS_clusters_e12))
row.names(mark_stats_e12) <-as.character(markersGS_clusters_e12@elementMetadata$name)

top_markers <- na.omit(unique(as.character(sapply(markersGS_e12, function(x) as.character(x$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|^Olfr", x$name)==F][1:10])))))

mark_stats_e12.plot <- mark_stats_e12[top_markers,]

## Cap to 2 /-1
mark_stats_e12.plot[mark_stats_e12.plot > 2] <- 2
mark_stats_e12.plot[mark_stats_e12.plot < (-1)] <- (-1)

print(draw(Heatmap(matrix = mark_stats_e12.plot,
        cluster_rows = T,
        cluster_columns = T,
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_row_names = T,
        show_column_names = T,
        col=viridis(100,option = "B"),
        border = TRUE,
        heatmap_legend_param = list(direction = "vertical", title = "Z-score GS")), heatmap_legend_side = "right"))
```

```{r, fig.width=5, fig.height=5}
lapply(clust_e12, function(c) {
  print(c)
  inspect_clust(c, markers = markersGS_e12)
})
```

###e13-e15

```{r}
clust_e13e15 <- names(which(table(proj$Clusters_within_Group1_astroglia[proj$Timepoint %in% c("e13","e15")]) > 50))

markersGS_clusters_e13e15 <-getMarkerFeatures(proj,
                  groupBy = paste0("Clusters_within_", subgroup),
                  useGroups = clust_e13e15,
                  bgdGroups = clust_e13e15,
                  useMatrix = "GeneScoreMatrix",
                  threads = 10,
                  scaleTo = 10^4)
markersGS_e13e15 <- getMarkers(markersGS_clusters_e13e15, cutOff = "FDR < 0.01 & Log2FC > 0.5")
```

```{r, fig.height=12, fig.width=5}
mark_stats_e13e15 <- as.matrix(assay(markersGS_clusters_e13e15))
row.names(mark_stats_e13e15) <-as.character(markersGS_clusters_e13e15@elementMetadata$name)

top_markers <- na.omit(unique(as.character(sapply(markersGS_e13e15, function(x) as.character(x$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|^Olfr", x$name)==F][1:8])))))

mark_stats_e13e15.plot <- mark_stats_e13e15[top_markers,]

## Cap to 2 /-1
mark_stats_e13e15.plot[mark_stats_e13e15.plot > 2] <- 2
mark_stats_e13e15.plot[mark_stats_e13e15.plot < (-1)] <- (-1)

print(draw(Heatmap(matrix = mark_stats_e13e15.plot,
        cluster_rows = T,
        cluster_columns = T,
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_row_names = T,
        show_column_names = T,
        col=viridis(100,option = "B"),
        border = TRUE,
        heatmap_legend_param = list(direction = "vertical", title = "Z-score GS")), heatmap_legend_side = "right"))
```

```{r, fig.width=5, fig.height=5}
lapply(clust_e13e15, function(c) {
  print(c)
  inspect_clust(c, markers = markersGS_e13e15)
})
```

###e17-P0

```{r}
clust_e17P0 <- names(which(table(proj$Clusters_within_Group1_astroglia[proj$Timepoint %in% c("e17","P0")]) > 50))

markersGS_clusters_e17P0 <-getMarkerFeatures(proj,
                  groupBy = paste0("Clusters_within_", subgroup),
                  useGroups = clust_e17P0,
                  bgdGroups = clust_e17P0,
                  useMatrix = "GeneScoreMatrix",
                  threads = 10,
                  scaleTo = 10^4)
markersGS_e17P0 <- getMarkers(markersGS_clusters_e17P0, cutOff = "FDR < 0.01 & Log2FC > 0.5")
```

```{r, fig.height=10, fig.width=5}
mark_stats_e17P0 <- as.matrix(assay(markersGS_clusters_e17P0))
row.names(mark_stats_e17P0) <-as.character(markersGS_clusters_e17P0@elementMetadata$name)

top_markers <- na.omit(unique(as.character(sapply(markersGS_e17P0, function(x) as.character(x$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|^Olfr", x$name)==F][1:10])))))

mark_stats_e17P0.plot <- mark_stats_e17P0[top_markers,]

## Cap to 2 /-1
mark_stats_e17P0.plot[mark_stats_e17P0.plot > 2] <- 2
mark_stats_e17P0.plot[mark_stats_e17P0.plot < (-1)] <- (-1)

print(draw(Heatmap(matrix = mark_stats_e17P0.plot,
        cluster_rows = T,
        cluster_columns = T,
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_row_names = T,
        show_column_names = T,
        col=viridis(100,option = "B"),
        border = TRUE,
        heatmap_legend_param = list(direction = "vertical", title = "Z-score GS")), heatmap_legend_side = "right"))
```

```{r, fig.width=5, fig.height=5}
lapply(clust_e17P0, function(c) {
  print(c)
  inspect_clust(c, markers = markersGS_e17P0)
})
```

###P4-P63

```{r}
clust_P4P63 <- names(which(table(proj$Clusters_within_Group1_astroglia[proj$Timepoint %in% c("P04","P07", "P14", "P63")]) > 50))

markersGS_clusters_P4P63 <-getMarkerFeatures(proj,
                  groupBy = paste0("Clusters_within_", subgroup),
                  useGroups = clust_P4P63,
                  bgdGroups = clust_P4P63,
                  useMatrix = "GeneScoreMatrix",
                  threads = 10,
                  scaleTo = 10^4)
markersGS_P4P63 <- getMarkers(markersGS_clusters_P4P63, cutOff = "FDR < 0.01 & Log2FC > 0.5")
```

```{r, fig.height=10, fig.width=5}
mark_stats_P4P63 <- as.matrix(assay(markersGS_clusters_P4P63))
row.names(mark_stats_P4P63) <-as.character(markersGS_clusters_P4P63@elementMetadata$name)

top_markers <- na.omit(unique(as.character(sapply(markersGS_P4P63, function(x) as.character(x$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|^Olfr", x$name)==F][1:10])))))

mark_stats_P4P63.plot <- mark_stats_P4P63[top_markers,]

## Cap to 2 /-1
mark_stats_P4P63.plot[mark_stats_P4P63.plot > 2] <- 2
mark_stats_P4P63.plot[mark_stats_P4P63.plot < (-1)] <- (-1)

print(draw(Heatmap(matrix = mark_stats_P4P63.plot,
        cluster_rows = T,
        cluster_columns = T,
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_row_names = T,
        show_column_names = T,
        col=viridis(100,option = "B"),
        border = TRUE,
        heatmap_legend_param = list(direction = "vertical", title = "Z-score GS")), heatmap_legend_side = "right"))
```

```{r, fig.width=5, fig.height=5}
lapply(clust_P4P63, function(c) {
  print(c)
  inspect_clust(c, markers = markersGS_P4P63)
})
```

Save markers

```{r}
saveRDS(markersGS_clusters_e10, "~/Mouse_Cereb/005_reprocess_by_cellType_broad/Group1_astroglia/Mouse_Cereb_subclusters_Group1_astroglia_res1.5_markerGenes_e10only.rds")
saveRDS(markersGS_clusters_e11, "~/Mouse_Cereb/005_reprocess_by_cellType_broad/Group1_astroglia/Mouse_Cereb_subclusters_Group1_astroglia_res1.5_markerGenes_e11only.rds")
saveRDS(markersGS_clusters_e12, "~/Mouse_Cereb/005_reprocess_by_cellType_broad/Group1_astroglia/Mouse_Cereb_subclusters_Group1_astroglia_res1.5_markerGenes_e12only.rds")
saveRDS(markersGS_clusters_e13e15, "~/Mouse_Cereb/005_reprocess_by_cellType_broad/Group1_astroglia/Mouse_Cereb_subclusters_Group1_astroglia_res1.5_markerGenes_e13e15only.rds")
saveRDS(markersGS_clusters_e17P0, "~/Mouse_Cereb/005_reprocess_by_cellType_broad/Group1_astroglia/Mouse_Cereb_subclusters_Group1_astroglia_res1.5_markerGenes_e17P0only.rds")
saveRDS(markersGS_clusters_P4P63, "~/Mouse_Cereb/005_reprocess_by_cellType_broad/Group1_astroglia/Mouse_Cereb_subclusters_Group1_astroglia_res1.5_markerGenes_P4P63only.rds")
```

Additionaly, we want to recluster C3, which seems to be a mixture of Gsx1+ bipotent progenitors and RL cells from e12.

```{r}
c3 <- subsetCells(proj,cellNames = proj$cellNames[proj$Clusters_within_Group1_astroglia=="C3"])

length(c3$cellNames)
table(c3$Timepoint)
```

Now we cluster again (using the previous LSI)

```{r, fig.width=6, fig.height=6}
c3 <- addClusters(input = c3,
                    name = "Clusters_within_Group1_astroglia_C3_sub",
                    reducedDims = paste0("IterativeLSI_", subgroup),
                    method = "Seurat",
                    force = T,
                    resolution=0.5,
                    corCutOff = 0.75,
                    scaleDims = FALSE,
                    seed = 1)

plotEmbedding(ArchRProj = c3, colorBy = "cellColData", name = "Clusters_within_Group1_astroglia_C3_sub", embedding = paste0("UMAP_within_", subgroup), rastr = T)
```

Let's call marker genes here:

```{r}
markersGS_c3_obj <-getMarkerFeatures(c3,
                  groupBy = "Clusters_within_Group1_astroglia_C3_sub",
                  useMatrix = "GeneScoreMatrix",
                  threads = 10,
                  scaleTo = 10^4)
markersGS_c3 <- getMarkers(markersGS_c3_obj, cutOff = "FDR < 0.05 & Log2FC > 0.5")
```

```{r, fig.height=10, fig.width=5}
mark_stats_c3 <- as.matrix(assay(markersGS_c3_obj))
row.names(mark_stats_c3) <-as.character(markersGS_c3_obj@elementMetadata$name)

top_markers <- na.omit(unique(as.character(sapply(markersGS_c3, function(x) as.character(x$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|^Olfr", x$name)==F][1:10])))))

mark_stats_c3.plot <- mark_stats_c3[top_markers,]

## Cap to 2 /-1
mark_stats_c3.plot[mark_stats_c3.plot > 2] <- 2
mark_stats_c3.plot[mark_stats_c3.plot < (-1)] <- (-1)

print(draw(Heatmap(matrix = mark_stats_c3.plot,
        cluster_rows = T,
        cluster_columns = T,
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_row_names = T,
        show_column_names = T,
        col=viridis(100,option = "B"),
        border = TRUE,
        heatmap_legend_param = list(direction = "vertical", title = "Z-score GS")), heatmap_legend_side = "right"))
```

C1: RL
C2: Progenitor
C3: Bipotent

```{r}
table(c3$Clusters_within_Group1_astroglia_C3_sub)
```

Let's update the clusters in the merged object

```{r}
proj$Clusters_within_Group1_astroglia_round1 <- proj$Clusters_within_Group1_astroglia

c3_clust <- as.data.frame(getCellColData(proj, select = c("cellNames_short","Sample_short","Clusters_within_Group1_astroglia_round1"))) %>%
  left_join(as.data.frame(getCellColData(c3, c("cellNames_short", "Clusters_within_Group1_astroglia_C3_sub"))))

sum(c3_clust$cellNames_short == proj$cellNames_short)
length(proj$cellNames_short)

c3_clust$Clusters_within_Group1_astroglia <- paste(c3_clust$Clusters_within_Group1_astroglia_round1, c3_clust$Clusters_within_Group1_astroglia_C3_sub, sep = ".")
c3_clust$Clusters_within_Group1_astroglia <- gsub(".NA", "", c3_clust$Clusters_within_Group1_astroglia)

table(c3_clust$Clusters_within_Group1_astroglia)
            
proj$Clusters_within_Group1_astroglia <- c3_clust$Clusters_within_Group1_astroglia
```

Saving relevant files:

```{r}
saveArchRProject(proj)
saveRDS(markersGS_clusters, paste0("Mouse_Cereb_subclusters_", subgroup, "_res1.5_markerGenes.rds"))
```


```{r}
sessionInfo()
```

