---
title: "Mouse Cerebellum 004 - Reprocess subgroup"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ArchR)
  library(tidyverse)
  library(RColorBrewer)
  library(ComplexHeatmap)
  library(viridis)
})
```

Group to be reprocessed:

```{r}
subgroup <- "Group2_earlyNuclei"

print(paste0("Reprocessing subgroup: ", subgroup))
```


Initialising ArchR

```{r}
addArchRThreads(threads = 20) 

## Working directory. This is where the arrow files will be generated and stored.
setwd(paste0("~/Mouse_Cereb/005_reprocess_by_cellType_broad/", subgroup))

## Genome and annotation
addArchRGenome("mm10")
```

```{r}
inputFiles <- list.files("fragments/", "_fragments.tsv.gz$")

inputFiles
names(inputFiles) <- sapply(inputFiles, function(x) gsub("_fragments.tsv.gz", "", x))
```

Now we can create our arrow files. We use the same QC metrics as before (TSS >=3, nFrags >= 5000) so all barcodes in the fragment files should pass QC:

```{r}
if (dir.exists("001_barcode_qc")==F){
  dir.create("001_barcode_qc")
}

ArrowFiles <- createArrowFiles(
  inputFiles = paste0("fragments/",inputFiles),
  sampleNames = names(inputFiles),
  filterTSS = 3, 
  filterFrags = 5000, 
  addTileMat = TRUE,
  addGeneScoreMat = TRUE,
  QCDir = "001_barcode_qc",
  promoterRegion = c(2000,100),
  removeFilteredCells = T,
  minFrags = 2500, 
  force = T
)
```

No need to calculate doublet scores or filter for doublets, these have already been removed from the merged dataset.

We can directly initialise the project.

```{r}
proj <- ArchRProject(
  ArrowFiles = ArrowFiles, 
  outputDirectory = paste0("proj_", subgroup),
  copyArrows = TRUE #This is recommened so that you maintain an unaltered copy for later usage.
)

## Also adding more metadata
proj$cellNames_short <- sapply(getCellNames(proj), function(x) {
  paste(str_extract(x, "sa[0-9]*"), str_split(x, pattern = fixed("#"), simplify = T)[2], sep="_")
})

proj$Sample_short <- sapply(proj$Sample, function(x) {
  x <- paste(str_split(x, pattern = fixed("_"), simplify = T)[c(4, 1, 5)], collapse="_")
  x <- gsub("P4", "P04", x)
  x <- gsub("P7", "P07", x)
})

proj$Timepoint <- sapply(proj$Sample_short, function(x) str_split(x, pattern = fixed("_"), simplify = T)[1])
proj$Lib <- sapply(proj$Sample_short, function(x) str_split(x, pattern = fixed("_"), simplify = T)[2])
proj$Sex <- sapply(proj$Sample_short, function(x) str_split(x, pattern = fixed("_"), simplify = T)[3])
```

Adding iterative LSI. Here we are analysing a more homogeneous set of cells compared to the entire dataset. Thus, we can reduce the number of dimensions and features to be used.

```{r}
proj <- addIterativeLSI(ArchRProj = proj,
                        useMatrix = "TileMatrix",
                        name = paste0("IterativeLSI_", subgroup),
                        iterations=5,
                        clusterParams = list(
                          resolution = c(0.1, 0.2, 0.4, 0.8), 
                          sampleCells = 10000, 
                          n.start = 10
                        ), 
                        varFeatures = 50000,
                        dimsToUse = 1:50,
                        totalFeatures = 200000,
                        seed = 1,
                        LSIMethod = 1,
                        scaleDims = FALSE,
                        corCutOff = 0.75,
                        excludeChr = c("chrX", "chrY", "chrMT"),
                        binarize = T,
                        force = T)
```

Adding clusters:

```{r}
proj <- addClusters(input = proj,
                    name = paste0("Clusters_within_", subgroup, "_res2.5"),
                    reducedDims = paste0("IterativeLSI_", subgroup),
                    method = "Seurat",
                    force = T,
                    resolution=2.5,
                    corCutOff = 0.75,
                    scaleDims = FALSE,
                    seed = 1)
```

And a UMAP. Here we are looking for substructure so we can reduce the number of neighbors:

```{r}
proj <- addUMAP(ArchRProj = proj,
                name = paste0("UMAP_within_", subgroup),
                reducedDims = paste0("IterativeLSI_", subgroup),
                minDist = 0.1,
                metric = "cosine",
                nNeighbors = 10,
                force = T,
                seed = 1,
                scaleDims = F,
                corCutOff = 0.75)
```

Plotting

```{r, fig.width=8, fig.height=8}
p1 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Sample_short", embedding = paste0("UMAP_within_", subgroup), pal = rev(colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample)))), rastr = T, labelMeans = F)
p2 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = paste0("Clusters_within_", subgroup), embedding = paste0("UMAP_within_", subgroup), rastr = T)

ggAlignPlots(p1, p2, type = "h")
```

```{r, fig.width=5, fig.height=5}
interesting_genes <- c("Notch1","Sox2","Fabp7","Cdon","Atoh1","Slc17a6", "Kirrel2","Gad2", "Meis1","Meis2", "Neurod1","Neurod2","Neurod6","Nr2f1","Lmo3","Lmx1a", "Lmx1b", "Pax5", "Pax8", "Tfap2b","Isl1","Isl2","Lef1","Pou6f2","Tlx3", "Evx1", "Evx2", "Sox14", "Skor2", "Lhx5")

interesting_genes <- interesting_genes[interesting_genes %in% proj@geneAnnotation$genes$symbol]

p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name = interesting_genes, 
    embedding = paste0("UMAP_within_", subgroup),
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)

p
```

Finding marker genes between the clusters

```{r}
markersGS_clusters <-getMarkerFeatures(proj,
                  groupBy = paste0("Clusters_within_", subgroup, "_res2.5"),
                  useGroups = names(which(table(getCellColData(proj, paste0("Clusters_within_", subgroup, "_res2.5"), drop = T)) > 50)),
                  useMatrix = "GeneScoreMatrix",
                  threads = 10,
                  scaleTo = 10^4)
markersGS <- getMarkers(markersGS_clusters, cutOff = "FDR < 0.01 & Log2FC > 0.5")
```

```{r, fig.height=12, fig.width=7}
mark_stats <- as.matrix(assay(markersGS_clusters))
row.names(mark_stats) <-as.character(markersGS_clusters@elementMetadata$name)

top_markers <- na.omit(unique(as.character(sapply(markersGS, function(x) as.character(x$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|^Olfr", x$name)==F][1:5])))))

mark_stats.plot <- mark_stats[top_markers,]

## Cap to 2 /-1
mark_stats.plot[mark_stats.plot > 2] <- 2
mark_stats.plot[mark_stats.plot < (-1)] <- (-1)

draw(Heatmap(matrix = mark_stats.plot,
        cluster_rows = T,
        cluster_columns = T,
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_row_names = T,
        show_column_names = T,
        col=viridis(100,option = "B"),
        border = TRUE,
        heatmap_legend_param = list(direction = "vertical", title = "Z-score GS")), heatmap_legend_side = "right")
```

Let's have a closer look into these clusters:

```{r}
prev_cells <- read_tsv("~/Mouse_Cerebellum/003_process_timepoints/005_annotate_cellTypes_all/Mouse_merged_seurat_cellTypes_allAssays_summary.txt")

inspect_clust <- function(clust,
                          archrproj=proj,
                          slot=paste0("Clusters_within_", subgroup, "_res2.5"),
                          markers=markersGS,
                          umap=paste0("UMAP_within_", subgroup)) {
  ## Gettting top 20 markers, excluding LncRNAs and miRNAs with non-informative names
  m <- markers[[clust]]$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|^Olfr", markers[[clust]]$name)==F][1:10]
  m <- na.omit(as.character(m))
  print("Top 10 markers")
  print(m)
  print("--------------------")
  ## Comparing to our previous annotation - cell type
  ## Index of cells belonging to the cluster
  i <- which(getCellColData(archrproj, slot,drop = T)==clust)
  print("Comparison to previous annotation")
  print(sort(table(prev_cells$cell_type[prev_cells$cell %in% archrproj$cellNames_short[i]]), decreasing = T))
  print("--------------------")
  ## Timepoint distribution
  print("Timepoint distribution")
  print(sort(table(archrproj$Clusters_within_Group2_earlyNuclei[i]), decreasing = T))
  print("--------------------")
  ## Previous cluster
  print(sort(table(archrproj$Timepoint[i]), decreasing = T))
  ## Plotting top genes
  
  if(length(m) > 0) {
    print("Plotting top marker genes")
      p <- plotEmbedding(
    ArchRProj = archrproj, 
    colorBy = "GeneScoreMatrix", 
    name =m, 
    embedding = umap,
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL,
  )
  print(p)
  } else {
    print("No marker genes")
  }

  print("--------------------")
}

lapply(sort(unique(archrproj$Clusters_within_Group2_earlyNuclei)), inspect_clust)
```

Saving relevant files:

```{r}
saveArchRProject(proj)
saveRDS(markersGS_clusters, paste0("Mouse_Cereb_subclusters_", subgroup, "_res1.5_markerGenes.rds"))
```


```{r}
sessionInfo()
```

