---
title: "Mouse Cerebellum 005 - Final cell type annotation"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ArchR)
  library(tidyverse)
  library(RColorBrewer)
  library(viridis)
  library(ComplexHeatmap)
  library(data.table)
})
```


Cell clusters

```{r}
proj_merged <- loadArchRProject("~/Mouse_Cereb/proj2_finalCells/")

proj_astroglia <- loadArchRProject("~/Mouse_Cereb/005_reprocess_by_cellType_broad/Group1_astroglia/proj_Group1_astroglia_subC3//")

proj_earlyNucs <- loadArchRProject("~/Mouse_Cereb/005_reprocess_by_cellType_broad/Group2_earlyNuclei//proj_Group2_earlyNuclei/")

proj_e13_Nb <- loadArchRProject("~/Mouse_Cereb/005_reprocess_by_cellType_broad/Group3_e13_GABA_Mix/proj_Group3_e13_GABA_Mix/")
```

Clustering overview

```{r}
cluster_sum <- as.data.frame(getCellColData(proj_merged, select = c("cellNames_short","Sample_short","Timepoint", "Sex", "Lib","Clusters_finalNucSet_res1.5"))) %>%
  left_join(as.data.frame(getCellColData(proj_astroglia, c("cellNames_short", "Clusters_within_Group1_astroglia")))) %>%
    left_join(as.data.frame(getCellColData(proj_earlyNucs, c("cellNames_short", "Clusters_within_Group2_earlyNuclei_res2.5")))) %>%
    left_join(as.data.frame(getCellColData(proj_e13_Nb, c("cellNames_short", "Clusters_within_Group3_e13_GABA_Mix"))))

table(cluster_sum$Clusters_finalNucSet_res1.5[cluster_sum$Clusters_within_Group2_earlyNuclei_res2.5=="C25"])
table(cluster_sum$Clusters_finalNucSet_res1.5[cluster_sum$Clusters_within_Group2_earlyNuclei_res2.5=="C26"])
table(cluster_sum$Clusters_finalNucSet_res1.5[cluster_sum$Clusters_within_Group2_earlyNuclei_res2.5=="C27"])

sum(proj_merged$cellNames_short == cluster_sum$cellNames_short)

proj_merged$Clusters_within_Group1_astroglia <- cluster_sum$Clusters_within_Group1_astroglia
proj_merged$Clusters_within_Group2_earlyNuclei_res2.5 <- cluster_sum$Clusters_within_Group2_earlyNuclei_res2.5
proj_merged$Clusters_within_Group3_e13_GABA_Mix <- cluster_sum$Clusters_within_Group3_e13_GABA_Mix
```

```{r, fig.width=12, fig.height=8}
p1 <- plotEmbedding(ArchRProj = proj_merged, colorBy = "cellColData", name = "Sample_short", embedding = "UMAP_finalNucSet", pal = colorRampPalette(brewer.pal(11, "Spectral"))(22), rastr = T, labelMeans = F)

p2 <- plotEmbedding(ArchRProj = proj_merged, colorBy = "cellColData", name = "Clusters_finalNucSet_res1.5", embedding = "UMAP_finalNucSet", rastr = T)

p3 <- plotEmbedding(ArchRProj = proj_merged, colorBy = "cellColData", name = "Clusters_within_Group2_earlyNuclei_res2.5", embedding = "UMAP_finalNucSet", rastr = T)

p4 <- plotEmbedding(ArchRProj = proj_merged, colorBy = "cellColData", name = "Clusters_within_Group1_astroglia", embedding = "UMAP_finalNucSet", rastr = T)

ggAlignPlots(p1, p2, p3, type = "h")
```


```{r,fig.width=8, fig.height=8}
p3
```

```{r,fig.width=8, fig.height=8}
plotEmbedding(ArchRProj = proj_merged, colorBy = "cellColData", name = "Clusters_within_Group2_earlyNuclei_res2.5", embedding = "UMAP_finalNucSet", rastr = T, labelMeans = F)
```


```{r,fig.width=12, fig.height=12}
p4
```

Cluster annotations

```{r}
anno_merged <- read_tsv("~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cerebellum_atac_merged_C47_anno.txt") %>% dplyr::rename(Cell_type_C47=Cell_type)

anno_astroglia <- read_tsv("~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cerebellum_atac_Group1_astroglia_anno.txt") %>% dplyr::rename(Cell_type_group1=Cell_type)

anno_earlyNucs <- read_tsv("~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cerebellum_atac_Group2_earlyNucs_anno.txt") %>% dplyr::rename(Cell_type_group2=Cell_type)

anno_e13_Nb <- read_tsv("~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cerebellum_atac_Group3_e13_Nbs_anno.txt") %>% dplyr::rename(Cell_type_group3=Cell_type)
```

General strategy

1. If only merged (no subclustering) -> keep merged
2. If astroglia: get cell type annotation from astroglia
3. If early Nucs: 
    If C15, C16:
      If anno-merged not mixed (ex. PC, interneuron) -> keep merged
      If anno-merged mixed (ex. C35-C39) -> keep early Nucs (Mixture_early_nuclei)
    If not C15, C16: keep early Nucs
4. If e13 Nb:
    If merged == C32: keep e13 Nb
    If merged != C32: keep merged

```{r}
cluster_sum <- left_join(cluster_sum, anno_merged) %>%
  left_join(anno_astroglia) %>%
  left_join(anno_earlyNucs) %>%
  left_join(anno_e13_Nb)

cluster_sum$Cell_type <- sapply(1:nrow(cluster_sum), function(x) {
  ifelse(is.na(cluster_sum$Cell_type_group1[x]) & is.na(cluster_sum$Cell_type_group2[x]) & is.na(cluster_sum$Cell_type_group3[x]), 
         yes = cluster_sum$Cell_type_C47[x],
         no = ifelse(is.na(cluster_sum$Cell_type_group1[x]) ==F ,
                     yes = cluster_sum$Cell_type_group1[x],
                     no = ifelse(is.na(cluster_sum$Cell_type_group2[x]) ==F ,
                                 yes = ifelse(cluster_sum$Clusters_within_Group2_earlyNuclei_res2.5[x] %in% c("C15", "C16"), 
                                              yes = ifelse(cluster_sum$Clusters_finalNucSet_res1.5[x] %in% c("C23", "C35", "C36", "C37", "C38", "C39"),
                                                           yes = cluster_sum$Cell_type_group2[x], 
                                                           no = cluster_sum$Cell_type_C47[x]),
                                              no = cluster_sum$Cell_type_group2[x]),
                                 no = ifelse(is.na(cluster_sum$Cell_type_group3[x]) ==F,
                                             yes = ifelse(cluster_sum$Clusters_finalNucSet_res1.5[x] == "C32", 
                                                          yes = cluster_sum$Cell_type_group3[x], 
                                                          no = cluster_sum$Cell_type_C47[x])))))
})

sort(table(cluster_sum$Cell_type), decreasing = T)

cluster_sum$Cell_type[cluster_sum$Cell_type=="Isthmic_parabrachial"] <- "Other"
cluster_sum$Cell_type[cluster_sum$Cell_type=="Mix_e13_Nbs"] <- "Other"

sum(is.na(cluster_sum$Cell_type))

proj_merged$Cell_type <- cluster_sum$Cell_type

sort(table(cluster_sum$Cell_type), decreasing = T)
```

```{r, fig.width=12,fig.height=12}
plotEmbedding(ArchRProj = proj_merged, colorBy = "cellColData", name = "Sample_short", embedding = "UMAP_finalNucSet", pal = colorRampPalette(brewer.pal(11, "Spectral"))(22), rastr = T, labelMeans = F)
plotEmbedding(ArchRProj = proj_merged, colorBy = "cellColData", name = "Cell_type", embedding = "UMAP_finalNucSet", rastr = T, labelMeans = T)
```

```{r, fig.width=12,fig.height=12}
metad <- as.data.frame(getCellColData(proj_merged))

row.names(metad) <- proj_merged$cellNames

umap <- getEmbedding(proj_merged,"UMAP_finalNucSet")
row.names(umap) <- proj_merged$cellNames
colnames(umap) <- c("UMAP1", "UMAP2")

all.df <- rownames_to_column(metad, "cell") %>%
  left_join(rownames_to_column(umap, "cell"))

labels.df <- group_by(all.df, Cell_type) %>%
  summarise(UMAP1=mean(UMAP1), 
            UMAP2=mean(UMAP2))


ggplot(all.df, aes(x=UMAP1, y=UMAP2, color=Cell_type)) +
  geom_point(size=0.05, alpha=0.7) +
  geom_label(data=labels.df, aes(label=Cell_type), position = position_jitterdodge(seed=1, jitter.width = 0.5, jitter.height = 1)) +
  theme_classic() +
  theme(legend.position = "none")
```


Let's check individual timepoints

```{r, fig.width=6, fig.height=6}
timepoints <- c("e10", "e11", "e12", "e13", "e15", "e17", "P0", "P4", "P7", "P14", "P63")
proj_byT <- lapply(timepoints, function(t) {
  proj_t <- loadArchRProject(paste0("~/Mouse_Cereb/010_byStage/proj_", t))
  t_sum <- as.data.frame(
    getCellColData(proj_t,
                   select = c("cellNames_short","Sample_short","Timepoint", "Sex", "Lib"))) %>%
  left_join(cluster_sum)
  proj_t$Cell_type <- t_sum$Cell_type
  print(paste0("Cell type abundance for timepoint", t))
  print(sort(table(proj_t$Cell_type), decreasing = T))
  print(plotEmbedding(ArchRProj = proj_t, colorBy = "cellColData", name = "Cell_type", embedding = paste0("UMAP_within_", t), rastr = T, labelMeans = T))
  print(plotEmbedding(ArchRProj = proj_t, colorBy = "cellColData", name = "DoubletEnrichment", embedding = paste0("UMAP_within_", t), rastr = T))
  print(plotEmbedding(ArchRProj = proj_t, colorBy = "cellColData", name = paste0("Clusters_within_", t), embedding = paste0("UMAP_within_", t), rastr = T, labelMeans = T))
  return(proj_t)
})

names(proj_byT) <- timepoints
```

```{r, fig.width=8, fig.height=8}
plotEmbedding(ArchRProj = proj_merged, colorBy = "cellColData", name = "DoubletEnrichment", embedding = "UMAP_finalNucSet", rastr = T)
```

The following stage-specific clusters appear to be problematic:

e11_C13
e12_C14
e12_C15
P14_C12

We should label them as put_doublets

```{r}
## e11
proj_e11 <- loadArchRProject("~/Mouse_Cereb/010_byStage/proj_e11/")

plotEmbedding(ArchRProj = proj_merged, highlightCells = proj_e11$cellNames[proj_e11$Clusters_within_e11=="C13"],embedding = "UMAP_finalNucSet", rastr = T, labelMeans = F )
sort(table(proj_merged$Cell_type[proj_merged$cellNames %in% proj_e11$cellNames[proj_e11$Clusters_within_e11=="C13"]]), decreasing = T)
table(proj_merged$Sample[proj_merged$cellNames %in% proj_e11$cellNames[proj_e11$Clusters_within_e11=="C13"]])

## e12
proj_e12 <- loadArchRProject("~/Mouse_Cereb/010_byStage/proj_e12/")

plotEmbedding(ArchRProj = proj_merged, highlightCells = proj_e12$cellNames[proj_e12$Clusters_within_e12=="C14"],embedding = "UMAP_finalNucSet", rastr = T, labelMeans = F )
sort(table(proj_merged$Cell_type[proj_merged$cellNames %in% proj_e12$cellNames[proj_e12$Clusters_within_e12=="C14"]]), decreasing = T)
table(proj_merged$Sample[proj_merged$cellNames %in% proj_e12$cellNames[proj_e12$Clusters_within_e12=="C14"]])

plotEmbedding(ArchRProj = proj_merged, highlightCells = proj_e12$cellNames[proj_e12$Clusters_within_e12=="C15"],embedding = "UMAP_finalNucSet", rastr = T, labelMeans = F )
sort(table(proj_merged$Cell_type[proj_merged$cellNames %in% proj_e12$cellNames[proj_e12$Clusters_within_e12=="C15"]]), decreasing = T)
table(proj_merged$Sample[proj_merged$cellNames %in% proj_e12$cellNames[proj_e12$Clusters_within_e12=="C15"]])

## P14
proj_P14 <- loadArchRProject("~/Mouse_Cereb/010_byStage/proj_P14/")

plotEmbedding(ArchRProj = proj_merged, highlightCells = proj_P14$cellNames[proj_P14$Clusters_within_P14=="C12"],embedding = "UMAP_finalNucSet", rastr = T, labelMeans = F )
sort(table(proj_merged$Cell_type[proj_merged$cellNames %in% proj_P14$cellNames[proj_P14$Clusters_within_P14=="C12"]]), decreasing = T)
table(proj_merged$Sample[proj_merged$cellNames %in% proj_P14$cellNames[proj_P14$Clusters_within_P14=="C12"]])
```

Updating the annotation for these cells as mixture/putative doublets

```{r}
proj_merged$Cell_type[proj_merged$cellNames %in% c(
  proj_e11$cellNames[proj_e11$Clusters_within_e11=="C13"],
  proj_e12$cellNames[proj_e12$Clusters_within_e12=="C14"],
  proj_e12$cellNames[proj_e12$Clusters_within_e12=="C15"],
  proj_P14$cellNames[proj_P14$Clusters_within_P14=="C12"]
)] <- "Other/mixture"

proj_merged$Cell_type[proj_merged$Cell_type=="Mixture_early_nuclei"] <- "Other/mixture"
proj_merged$Cell_type[proj_merged$Cell_type=="Other"] <- "Other/mixture"
proj_merged$Cell_type[proj_merged$Cell_type=="Other"] <- "Other/mixture"
```

```{r, fig.width=12,fig.height=12}
plotEmbedding(ArchRProj = proj_merged, colorBy = "cellColData", name = "Sample_short", embedding = "UMAP_finalNucSet", pal = colorRampPalette(brewer.pal(11, "Spectral"))(22), rastr = T, labelMeans = F)
plotEmbedding(ArchRProj = proj_merged, colorBy = "cellColData", name = "Cell_type", embedding = "UMAP_finalNucSet", rastr = T, labelMeans = T)
```

```{r, fig.width=12,fig.height=12}
metad <- as.data.frame(getCellColData(proj_merged))

row.names(metad) <- proj_merged$cellNames

umap <- getEmbedding(proj_merged,"UMAP_finalNucSet")
row.names(umap) <- proj_merged$cellNames
colnames(umap) <- c("UMAP1", "UMAP2")

all.df <- rownames_to_column(metad, "cell") %>%
  left_join(rownames_to_column(umap, "cell"))

labels.df <- group_by(all.df, Cell_type) %>%
  summarise(UMAP1=mean(UMAP1), 
            UMAP2=mean(UMAP2))


ggplot(all.df, aes(x=UMAP1, y=UMAP2, color=Cell_type)) +
  geom_point(size=0.05, alpha=0.7) +
  geom_label(data=labels.df, aes(label=Cell_type), position = position_jitterdodge(seed=1, jitter.width = 0.5, jitter.height = 2)) +
  theme_classic() +
  theme(legend.position = "none")
```

Export cell type annotation

```{r}
cell_anno <- getCellColData(proj_merged) %>%
  as.data.frame() %>%
  rownames_to_column("cell") %>%
  select(-Cell_type_C47, -Cell_type_broad_C47)

sort(table(cell_anno$Cell_type), decreasing = T)
```

We also need to group together in broad cell types and set up a color scheme:

```{r}
length(unique(cell_anno$Cell_type))
#write(sort(unique(cell_anno$Cell_type)), "~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cerebellum_atac_unique_cell_type_labels.txt")
```

```{r}
cell_cols <- read_tsv("~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cerebellum_atac_unique_cell_type_labels_andColors.txt")

cell_anno <- left_join(cell_anno, cell_cols)

cell_anno$Cell_type <- factor(cell_anno$Cell_type, levels = unique(cell_cols$Cell_type))
cell_anno$Cell_type_broad <- factor(cell_anno$Cell_type_broad, levels = unique(cell_cols$Cell_type_broad))
cell_anno$Cell_type_color <- factor(cell_anno$Cell_type_color, levels = unique(cell_cols$Cell_type_color))
cell_anno$Cell_type_broad_color <- factor(cell_anno$Cell_type_broad_color, levels = unique(cell_cols$Cell_type_broad_color))
```

Exporting annotation:

```{r}
write_tsv(cell_anno, "~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cerebellum_atac_finalCellType_annotation.txt")
saveRDS(cell_anno, "~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cerebellum_atac_finalCellType_annotation.rds")
```

```{r}
sessionInfo()
```

