---
title: "Mouse Cerebellum 006 - Peak calling"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ArchR)
  library(tidyverse)
  library(RColorBrewer)
  library(viridis)
  library(ComplexHeatmap)
})
```

We will use ArchR's approach for reproducible peak calling within groups (i.e. requiring a peak to be robustly active within a group across biological replicates).

Choosing the correct grouping is a difficult task with tradeoffs in every decision. 

If we aim for maximum granularity (clusters + subclusters) we will have too few cells to work with.
If we group together cell types, we might miss important developmental signal (ex. subclusters assigned to RL from e10, e11, e12 are still very different from each other).

For this reason we will work with our 47 clusters. This gives us a reasonable number of cells per cluster and still enough resolution to detect peaks active in different cell types/states.

First, let's create a copy of our project:

```{r}
addArchRThreads(threads = 30) 

setwd("~/Mouse_Cereb/")
proj_old <- loadArchRProject("proj2_finalCells")
saveArchRProject(proj_old, "proj2_backup", load = F)
```

Another important parameter we have to consider for peak calling is the optimal number of cells.

For that we should first have an idea of how many cells per replicate we have in each group

```{r}
clust_info <- getCellColData(proj_old) %>%
  as.data.frame()

clust_sum <- data.frame(clust=factor(names(sort(table(clust_info$Clusters_finalNucSet_res1.5), decreasing = T)),levels = names(sort(table(clust_info$Clusters_finalNucSet_res1.5), decreasing = T))),
                        counts=as.numeric(sort(table(clust_info$Clusters_finalNucSet_res1.5), decreasing = T)))

ggplot(data = clust_sum, aes(x=clust, y=counts)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle=45, hjust=0.9)) +
  geom_hline(yintercept = 600, color="red", lty="dashed")+
  geom_hline(yintercept = 2500, color="red", lty="dashed")

summary(as.numeric(table(clust_info$Clusters_finalNucSet_res1.5)))
```




```{r}
clust_info.sum <- group_by(clust_info,Clusters_finalNucSet_res1.5, Timepoint, Lib) %>%
  count() %>%
  ungroup() %>%
  group_by(Clusters_finalNucSet_res1.5) %>%
  summarise(nCells_total = sum(n),
            nCells_max=max(n),
            nCells_2nd=sort(n, decreasing = T)[2],
            lib_max=Lib[which.max(n)],
            lib_2nd=Lib[which(n==sort(n, decreasing = T)[2])[1]],
            timepoint_max=Timepoint[which.max(n)],
            timepoint_2nd=Timepoint[which(n==sort(n, decreasing = T)[2])[1]], 
            nReps=n_distinct(Lib), 
            nTimepoint=n_distinct(Timepoint)) %>%
  ungroup() %>%
  arrange(desc(nCells_max)) %>%
  mutate(cluster47=factor(Clusters_finalNucSet_res1.5, levels = Clusters_finalNucSet_res1.5))

summary(clust_info.sum$nCells_total)
summary(clust_info.sum$nCells_max)
summary(clust_info.sum$nReps)
```


```{r}
clust_info$Clusters_finalNucSet_res1.5 <- factor(clust_info$Clusters_finalNucSet_res1.5, levels = levels(clust_info.sum$cluster47))

ggplot(clust_info, aes(Clusters_finalNucSet_res1.5, fill=Sample_short)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.9), legend.position = "top")
```

```{r}
ggplot(clust_info.sum, aes(x=nCells_2nd, y=nReps, color=nCells_total)) +
  geom_point() +
  ylim(c(0,22)) +
  geom_hline(yintercept = 2, color="red", lty="dashed")
```

We see that smaller clusters typically contain cells from more samples, whereas bigger samples are often more timepoint specific.

We should constraint the amount of cells we get from clusters with a few replicates, whereas allowing ourselves to draw information from multiple samples for smaller clusters.

Creating pseudobulk replicates:

```{r}
proj <- addGroupCoverages(
  ArchRProj = proj_old,
  minCells = 100, ## minimum number of cells per replicate -> thus minimum number of cells per group will be 2 reps x 100 cells = 200 cells!
  maxCells = 1000, ## maximum number of cells per replicate
  maxFragments = 50*1e6, ## maximum number of fragments per cell group
  minReplicates = 2, ## minimum number of replicates per group
  maxReplicates = 10, ## maximum number of replicates per group
  sampleRatio = 0.8, ## fraction of cells that can be sampled from a replicate to contribute to other replicates (without replacement) -> this will only affect our three clusters with less than 200 cells!
  returnGroups = F,
  force = T,
  groupBy = "Clusters_finalNucSet_res1.5"
)
```

Now calling peaks:

```{r}
pathToMacs2 <- "/home/ioansarr/Software/bin/macs2"

proj <- addReproduciblePeakSet(
    ArchRProj = proj, 
    groupBy = "Clusters_finalNucSet_res1.5", 
    pathToMacs2 = pathToMacs2,
    reproducibility = "2", 
    peaksPerCell = 1000,
    maxPeaks = 200000,
    minCells = 40,
    excludeChr = c("chrY", "chrMT"), 
    method = "q",
    cutOff = 0.01, 
    extendSummits = 250,
    force = T
)
```

```{r}
peaks <- getPeakSet(proj)
head(peaks)
length(peaks)
```

Also getting the peak matrix:

```{r}
proj <- addPeakMatrix(proj,
                      ceiling = 5,
                      binarize = F,
                      force = T)

peak_mat <- getMatrixFromProject(proj, "PeakMatrix")
```

```{r}
saveArchRProject(proj, "proj3_peakCalling/")
saveRDS(peaks, "006_peaks/Mouse_Cerebellum_reproducible_peaks.rds")
saveRDS(peak_mat, "006_peaks/Mouse_Cerebellum_reproducible_peak_matrix_SE.rds")
```

```{r}
sessionInfo()
```
