---
title: "Mouse Cerebellum 008 - Peak co-accessibility"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ArchR)
  library(tidyverse)
  library(RColorBrewer)
  library(viridis)
  library(ComplexHeatmap)
})
```

First, let's create a copy of our project:

```{r}
addArchRThreads(threads = 30) 
setwd("~/Mouse_Cereb/")
proj <- loadArchRProject("proj4_chromVar/")
```

Now we can estimate peak co-accessibility through ArchR. This method is similar but not the same with Cicero (doesn't seem to penalise distance, which in this case is advantageous).
The main purpose of this analysis is to identify co-accessible peaks (not necessarily to link peaks to genes, for which we will take into consideration gene activity scores/imputed RNA-seq expression).

```{r}
proj <- addCoAccessibility(
    ArchRProj = proj,
    reducedDims = "IterativeLSI_finalNucSet", 
    dimsToUse = 1:ncol(proj@reducedDims$IterativeLSI_finalNucSet$matSVD),
    scaleDims = F,
    k = 50,
    knnIteration = 5000,
    overlapCutoff = 0.5,
    maxDist = 1e5,
    seed = 1, 
    log2Norm = T
)
```

Getting co-accessibility as a dataframe. Let's be very lenient with our correlation cutoff. We can always filter correlations later.
We are also interested in peak-peak linkages, so we should use the maximum resolution (1bp).

```{r}
cicero_conns_all <- getCoAccessibility(proj,
                   corCutOff = (-1),
                   resolution = 1, 
                   returnLoops = F)

cicero_conns_0.5 <- getCoAccessibility(proj,
                   corCutOff = 0.5,
                   resolution = 1, 
                   returnLoops = F)

```

Saving the data

```{r}
saveRDS(cicero_conns_all, "008_cicero/Mouse_Cerebellum_cicero_conns_all.rds")
saveRDS(cicero_conns_0.5, "008_cicero/Mouse_Cerebellum_cicero_conns_corrCutOff_0.5.rds")
saveArchRProject(proj)
```

```{r}
sessionInfo()
```

