---
title: "Mouse Cerebellum - Benchmarking peaks with external datasets - Cusanovich Cell 2018"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(Matrix)
  library(gridExtra)
})
```

```{r}
peaks <- read.table("~/Mouse_Cereb/006_peaks/Mouse_Cerebellum_reproducible_peaks_summary.txt", header = T, sep = "\t", stringsAsFactors = F)
```

Reading in the raw binary matrix

```{r}
cusanovich.data <- readRDS("~/Mouse_Cereb/006_peaks/002_external_data/Cusanovich_Cell_2018/atac_matrix.binary.qc_filtered.rds")

dim(cusanovich.data)

colnames(cusanovich.data)[1:10]
row.names(cusanovich.data)[1:10]
```

Reading in their peaks and intersecting with our dataset

```{r, eval=FALSE}
cusanovich.peaks <- data.frame(peak=row.names(cusanovich.data), stringsAsFactors = F) %>%
  separate(peak, into = c("chr", "start", "end"), remove = F, sep = "_") %>%
  mutate(start=as.numeric(start), end=as.numeric(end)) %>%
  select(chr, start, end, peak)

head(cusanovich.peaks)

table(cusanovich.peaks$chr)

peak_width <- cusanovich.peaks$end - cusanovich.peaks$start

hist(peak_width)
summary(peak_width)
```

First, we have to liftover to mm10, as they have used mm9 for their study!!!

```{r, eval=FALSE}
write.table(cusanovich.peaks, "~/Mouse_Cereb/006_peaks/002_external_data/Cusanovich_Cell_2018/peak_annotation/cusanovich_peaks_mm9.bed", col.names = F, row.names = F, sep = "\t", quote = F)
```

```{bash, eval=FALSE}
~/Software/bin/liftOver ~/Mouse_Cereb/006_peaks/002_external_data/Cusanovich_Cell_2018/peak_annotation/cusanovich_peaks_mm9.bed ~/Resources/liftover_chains/mm9ToMm10.over.chain.gz -minMatch=0.9 ~/Mouse_Cereb/006_peaks/002_external_data/Cusanovich_Cell_2018/peak_annotation/cusanovich_peaks_mm10.bed ~/Mouse_Cereb/006_peaks/002_external_data/Cusanovich_Cell_2018/peak_annotation/cusanovich_peaks_mm9_unlifted.bed
```

Reading in the liftovered files (and replacing the previous object):

```{r}
cusanovich.peaks <- read_tsv("~/Mouse_Cereb/006_peaks/002_external_data/Cusanovich_Cell_2018/peak_annotation/cusanovich_peaks_mm10.bed", col_names = c("chr", "start", "end", "peak"))

dim(cusanovich.peaks)
dim(cusanovich.data)

## Subsetting the matrix for the peaks that were successfully liftovered
cusanovich.data <- cusanovich.data[cusanovich.peaks$peak, ]
```

Overlapping with our peaks (based on the entire peak width). We will again limit ourselves to the robust peaks.

```{bash}
bedtools intersect -a ~/Mouse_Cereb/006_peaks/Mouse_Cerebellum_reproducible_peaks_robust.bed -b  ~/Mouse_Cereb/006_peaks/002_external_data/Cusanovich_Cell_2018/peak_annotation/cusanovich_peaks_mm10.bed -wa -wb > ~/Mouse_Cereb/006_peaks/002_external_data/Cusanovich_Cell_2018/peak_annotation/Mouse_Cerebellum_overlapping_cusanovich_peaks.txt
```

Let's read in the overlaps

X4: Mouse Cerebellum peaks
X8: Cusanovich peaks

```{r}
cerebellum_overlapping_cusanovich <- read_tsv("~/Mouse_Cereb/006_peaks/002_external_data/Cusanovich_Cell_2018/peak_annotation/Mouse_Cerebellum_overlapping_cusanovich_peaks.txt", col_names = F)
```

First, let's see whether the peaks found in our dataset are more active in cerebellum (and other brain cell types). 

For that we also need to read in the metadata for the barcodes:

```{r}
cusanovich_metadata <- read_tsv("~/Mouse_Cereb/006_peaks/002_external_data/Cusanovich_Cell_2018/cell_metadata.txt")

dim(cusanovich_metadata)
head(cusanovich_metadata)
table(cusanovich_metadata$tissue)
table(cusanovich_metadata$cell_label)
```

For each cell type and tissue, we can calculate a relative enrichment score of the peaks found in our annotation versus all peaks (counts_peaks_overlap/counts_peaks).

```{r}
inOurPeaks <- unique(cerebellum_overlapping_cusanovich$X8)

peak_enrichment <- data.frame(cell=colnames(cusanovich.data),
                              total=Matrix::colSums(cusanovich.data),
                              inOur=Matrix::colSums(cusanovich.data[inOurPeaks,]),
                              stringsAsFactors = F)

peak_enrichment$cer_enrichment <- peak_enrichment$inOur / peak_enrichment$total

peak_enrichment <- left_join(cusanovich_metadata, peak_enrichment)
```

```{r}
## Order to plot tissues
tissue_order <- group_by(peak_enrichment, tissue) %>%
  summarise(cer_enr_med=median(cer_enrichment)) %>%
  arrange(desc(cer_enr_med))

peak_enrichment$tissue <- factor(peak_enrichment$tissue, tissue_order$tissue)

p <- ggplot(peak_enrichment, aes(x=tissue, y=cer_enrichment, fill=tissue)) +
  geom_violin() +
  geom_boxplot(width=0.1, outlier.alpha = 0) +
  theme_classic() +
  ylab("Fraction of counts in cerebellum peaks") +
  xlab("Tissue (Cusanovich 2018)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust = 0.95), legend.position = "none")

p

pdf("~/Mouse_Cereb/Figures/003_peakAnnotation/Cusanovich_FractionOfReads_inCerebellumRobustPeaks_byTissue.pdf", width = 5, height = 4, useDingbats = F); print(p); dev.off()
```

Same for cell types:

```{r}
cell_label_order <- group_by(peak_enrichment, cell_label) %>%
  summarise(cer_enr_med=median(cer_enrichment)) %>%
  arrange(desc(cer_enr_med))

peak_enrichment$cell_label <- factor(peak_enrichment$cell_label, cell_label_order$cell_label)

ggplot(peak_enrichment, aes(x=cell_label, y=cer_enrichment, fill=cell_label)) +
  geom_violin() +
  geom_boxplot(width=0.1, outlier.alpha = 0) +
  theme_classic() +
  ylab("Fraction of counts in cerebellum peaks") +
  xlab("Cell type (Cusanovich 2018)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust = 0.95), legend.position = "none")


## Only plotting top-10 and bottom-10 cell types
p <- ggplot(filter(peak_enrichment, cell_label %in% cell_label_order$cell_label[c(1:10, (nrow(cell_label_order)-9): nrow(cell_label_order))]), aes(x=cell_label, y=cer_enrichment, fill=cell_label)) +
  geom_violin() +
  geom_boxplot(width=0.1, outlier.alpha = 0) +
  theme_classic() +
  geom_vline(xintercept = 10.5, color="indianred", lty="dashed") +
  ylab("Fraction of counts in cerebellum peaks") +
  xlab("Cell type (Cusanovich 2018)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust = 0.95), legend.position = "none")

p

pdf("~/Mouse_Cereb/Figures/003_peakAnnotation/Cusanovich_FractionOfReads_inCerebellumRobustPeaks_byCellType.pdf", width = 5, height = 4, useDingbats = F); print(p); dev.off()
```


```{r}
sessionInfo()
```

