---
title: "Opossum Cerebellum 001 - Barcode QC with ArchR"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ArchR)
  library(tidyverse)
  library(RColorBrewer)
})
```

Initialising ArchR

```{r}
addArchRThreads(threads = 40) 

## Working directory. This is where the arrow files will be generated and stored.
setwd("~/Opo_Cereb/")

## Genome and annotation
mondom5_genomeAnno <- readRDS("~/Opo_Cereb/000_prep_annotation/002_makeArchRanno/ArcR_monDom5_splitChr_genomeAnnotation.rds")
mondom5_geneAnno <- readRDS("~/Opo_Cereb/000_prep_annotation/002_makeArchRanno/ArcR_monDom5_splitChr_ens96andXLOCs_geneAnnotation.rds")
```

Reading in the fragment files and creating the respective arrow files

```{r}
samples <- list.files("~/sds/sd17d003/scATAC/Cerebellum/Opossum/", pattern = "^sa.*", include.dirs = T)

sample.info <- data.frame(sample=samples, s_split=samples,stringsAsFactors = F) %>%
  separate(s_split, into = c("lib", "species", "tissue", "timepoint", "sex"), sep = "_")

inputFiles <- sapply(samples, function(x) paste0("~/sds/sd17d003/scATAC/Cerebellum/Opossum/",x, "/cellranger/", x, "/outs/fragments.tsv.gz"))

length(samples)
print(samples)
```

Now creating the arrow files and filtering for high quality cells.

We will require high quality barcodes to have at least 5,000 fragments and a TSS enrichment of 2 (same fragments as mouse but TSS enrichment reduced as the opossum annotation is less precise).

```{r}
if (dir.exists("001_barcode_qc")==F){
  dir.create("001_barcode_qc")
}

ArrowFiles <- createArrowFiles(
  inputFiles = inputFiles,
  sampleNames = names(inputFiles),
  filterTSS = 2, 
  filterFrags = 5000,
  genomeAnnotation = mondom5_genomeAnno,
  geneAnnotation = mondom5_geneAnno, 
  addTileMat = TRUE,
  addGeneScoreMat = TRUE,
  QCDir = "001_barcode_qc",
  promoterRegion = c(2000,100),
  minFrags = 2500,
  force = T
)
```

```{r}
ArrowFiles <- list.files("~/Opo_Cereb/", pattern = ".arrow")
```

```{r}
if (dir.exists("002_doublets")==F){
  dir.create("002_doublets")
}

doubScores <- addDoubletScores(
  input = ArrowFiles,
  k = 10, #Refers to how many cells near a "pseudo-doublet" to count.
  knnMethod = "LSI", #Refers to the embedding to use for nearest neighbor search.
  LSIMethod = 1,
  dimsToUse = 1:50,
  nTrials = 10,
  scaleDims = F,
  LSIParams = list(seed = 1,
                   varFeatures = 50000,
                   excludeChr = c("chrX", "chrY", "chrM")),
  outDir = "002_doublets"
)
```

Now we can initialize our first project:

```{r}
proj <- ArchRProject(
  ArrowFiles = ArrowFiles,  
  genomeAnnotation = mondom5_genomeAnno,
  geneAnnotation = mondom5_geneAnno,
  outputDirectory = "proj1_init",
  copyArrows = TRUE #This is recommened so that you maintain an unaltered copy for later usage.
)

## Also adding more metadata
proj$cellNames_short <- sapply(getCellNames(proj), function(x) {
  paste(str_extract(x, "sa[0-9]*"), str_split(x, pattern = fixed("#"), simplify = T)[2], sep="_")
})

proj$Tissue <- sapply(proj$Sample, function(x) str_split(x, pattern = fixed("_"), simplify = T)[3])

proj$Timepoint <- sapply(proj$Sample, function(x) str_split(x, pattern = fixed("_"), simplify = T)[4])

proj$Lib <- sapply(proj$Sample, function(x) str_split(x, pattern = fixed("_"), simplify = T)[1])
proj$Sex <- sapply(proj$Sample, function(x) str_split(x, pattern = fixed("_"), simplify = T)[5])

proj$Sample_short <- paste(proj$Tissue, proj$Timepoint, proj$Lib, proj$Sex, sep = "_")
```

How many cells passed QC? Start keeping records for them:

```{r}
nCells(proj)
barcodes_init <- getCellColData(proj) %>%
  as.data.frame() %>%
  rownames_to_column("barcode")
```

TSS ridge plot

```{r}
p <- plotGroups(
    ArchRProj = proj, 
    groupBy = "Sample_short", 
    colorBy = "cellColData", 
    name = "TSSEnrichment",
    pal = rev(colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample_short)))),
    plotAs = "ridges"
   ) +
  ylab("Samples")

p
```

Fragment size distribution

```{r}
fragment_sizes <- plotFragmentSizes(ArchRProj = proj, returnDF = T)

head(fragment_sizes)
dim(fragment_sizes)

fragment_sizes <- as.data.frame(fragment_sizes) %>%
  dplyr::rename(Sample=sampleName) %>%
  left_join(unique(dplyr::select(as.data.frame(getCellColData(proj)), Sample, Sample_short)))

p <-ggplot(fragment_sizes, aes(x=fragmentSize, y=fragmentPercent, color=as.factor(Sample_short))) +
  geom_line() +
  theme_classic() +
  xlab("Fragment size") +
  ylab("Density") +
  theme(legend.position = "bottom")+
  scale_color_manual(values = colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample_short))), name="Sample")

p
```

TSS enrichment

```{r}
tssEnr <- plotTSSEnrichment(ArchRProj = proj, returnDF = T)

head(tssEnr)
dim(tssEnr)

tssEnr <- as.data.frame(tssEnr) %>%
  dplyr::rename(Sample=sampleName) %>%
  left_join(unique(dplyr::select(as.data.frame(getCellColData(proj)), Sample, Sample_short)))

p <-ggplot(tssEnr, aes(x=x, y=smoothValue, color=as.factor(Sample_short))) +
  geom_line() +
  theme_classic() +
  xlab("Distance from TSS (bp)") +
  ylab("Normalised signal") +
  theme(legend.position = "bottom")+
  scale_color_manual(values = colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample_short))), name="Sample")

p
```

Filtering putative doublets (pass 1).
Here we want to be relatively lenient, so we select a ratio of 1 (i.e. filtering top 5% of barcodes for a sample of 5,000 cells).

We are using Doublet enrichment as a cutoff.

```{r}
proj <- filterDoublets(ArchRProj = proj,
                       filterRatio = 1)
```

Updating these barcodes in our log file:

```{r}
nCells(proj)
barcodes_init$Doublet_1stPass <- !(barcodes_init$barcode %in% getCellNames(proj))
```

Now we can calculate our first iterative LSI. We are now extending our dimensions to 100 and the number of variable features to 100K.

```{r}
proj <- addIterativeLSI(ArchRProj = proj,
                        useMatrix = "TileMatrix",
                        name = "IterativeLSI_qc",
                        iterations=5,
                        clusterParams = list(
                          resolution = c(0.1, 0.2, 0.4, 0.8), 
                          sampleCells = 20000, 
                          n.start = 10
                        ), 
                        varFeatures = 100000,
                        dimsToUse = 1:100,
                        totalFeatures = 500000,
                        seed = 1,
                        LSIMethod = 1,
                        scaleDims = FALSE,
                        corCutOff = 0.75,
                        excludeChr = c("chrX", "chrY", "chrM"),
                        binarize = T,
                        force = T)
```

Clustering: The goal here is to identify clusters enriched for doublets/ low quality cells.
For this we should go for a relatively high resolution (3.0) for Seurat Louvain clustering.

```{r}
proj <- addClusters(input = proj,
                    name = "Clusters_qc",
                    reducedDims = "IterativeLSI_qc",
                    method = "Seurat",
                    force = T,
                    resolution=3.0,
                    corCutOff = 0.75,
                    scaleDims = FALSE,
                    maxClusters = 80,
                    seed = 1)

barcodes_init <- getCellColData(proj, select = "Clusters_qc") %>%
  as.data.frame() %>%
  rownames_to_column("barcode") %>%
  right_join(barcodes_init)
```

We can also run a UMAP to allow us to visualise the distribution of putative doublets/ low quality cells.

```{r}
proj <- addUMAP(ArchRProj = proj,
                name = "UMAP_qc",
                reducedDims = "IterativeLSI_qc",
                minDist = 0.2,
                metric = "cosine",
                nNeighbors = 25,
                force = T,
                seed = 1,
                scaleDims = F,
                corCutOff = 0.75)
```

Now we can have a look into the UMAP:

```{r, fig.width=8, fig.height=8}
p1 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Sample_short", embedding = "UMAP_qc", pal = rev(colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample_short)))), rastr = T, labelMeans = F)
p2 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Clusters_qc", embedding = "UMAP_qc", rastr = T)

ggAlignPlots(p1, p2, type = "h")
```

```{r}
if (dir.exists("Figures/001_QC")==F){
  dir.create("Figures/001_QC",recursive = T)
}

pdf("Figures/001_QC/Opossum_atac_LSI_pass1_tiles_bySample.pdf", width = 8, height = 8, useDingbats = F); print(p1);dev.off()

pdf("Figures/001_QC/Opossum_atac_LSI_pass1_tiles_byCluster.pdf", width = 8, height = 8, useDingbats = F); print(p2);dev.off()
```

Let's also plot some genes

```{r}
interesting_genes <- toupper(c("Atoh1", "Sox2", "Notch1","Cyp26b1","Foxp2", "Skor2", "Car8", "Cdon", "Gad2", "Sox14", "Tlx3", "Lmx1a", "Eomes", "Mgat5","Pax2", "Top2a", "Ptf1a", "Kirrel2", "Rora", "Rorb", "Slc17a6", "Meis2", "Aqp4", "Olig2", "Cbln3", "Fgf3", "Grin2b", "Neurod1", "Nhlh1", "Gdf10", "Fabp7", "Map3k1", "Gli2", "Esrrg", "Sorcs3", "Cdh22", "Etv1", "Junb", "Mef2a", "Foxp1", "Cdh9", "March11", "Pvalb", "Lhx2", "Pax5", "Zfhx3", "Zfhx4", "Fgf17", "Syndig1l","Gabra6", "Slc1a3", "Inpp5d", "Cx3cr1", "Mobp", "Sox10", "Lims2", "Slc4a1", "Angptl8", "Lmx1b", "Isl1", "Lef1"))

interesting_genes <- interesting_genes[interesting_genes %in% proj@geneAnnotation$genes$symbol]

p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name = interesting_genes, 
    embedding = "UMAP_qc",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)

p
```

We see some differences in the P21 samples (could be developmental differences or batch effects). We'll need to use Harmony to align these samples in order to call peaks.

But first, we can make sure we have removed all doublets (similar to our analysis on mouse):

```{r}
## Doublet score
p <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "DoubletScore", embedding = "UMAP_qc", rastr = T)
p


## Doublet enrichment
p <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "DoubletEnrichment", embedding = "UMAP_qc", rastr = T)
p

## nFragments
p <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "nFrags", embedding = "UMAP_qc", rastr = T)
p
```

```{r, fig.width=6, fig.height=6}
hist(proj$nFrags, breaks = 100, main = "Number of fragments")
abline(v = median(proj$nFrags), col="deepskyblue3", lty="dashed")

p <- ggPoint(x=proj$nFrags,
        y=proj$DoubletEnrichment,
        colorDensity = T,
        rastr = T) +
  geom_hline(yintercept = 4, col="red", lty="dashed") +
  geom_vline(xintercept = 45000, col="red", lty="dashed") +
  xlab("nFragments") +
  ylab("DoubletEnrichment")

p

```


```{r, fig.width=8, fig.height=8}
df <- getCellColData(proj) %>%
  as.data.frame()

median(df$nFrags)

ggplot(df, aes(log10(nFrags))) +
  geom_histogram(bins = 50) +
  facet_wrap(~Sample_short) +
  geom_vline(xintercept = log10(2.5*median(df$nFrags)), color="red")

doub_cutoffs <- group_by(df, Sample) %>%
  summarise(
    medFrags = median(nFrags),
    doubCutoff = 2.5 * median(nFrags)
  )

put_doublets <- proj$cellNames[sapply(1:length(proj$cellNames), function(i) proj$nFrags[i] > doub_cutoffs$doubCutoff[doub_cutoffs$Sample == proj$Sample[i]] | proj$DoubletEnrichment[i] > 4)]

barcodes_init$Doublet_2ndPass <- barcodes_init$barcode %in% put_doublets
```

```{r}
doub_enrich <- getCellColData(proj) %>%
  as.data.frame() %>%
  rownames_to_column("cellNames") %>%
  group_by(Clusters_qc) %>%
  summarise(Doub_freq=sum(cellNames %in% put_doublets)/length(cellNames)) %>%
  arrange(desc(Doub_freq)) %>%
  mutate(Clusters_qc=factor(Clusters_qc, levels = unique(Clusters_qc)))
  
p <- ggplot(doub_enrich, aes(x=Clusters_qc, y=Doub_freq)) +
  geom_bar(stat="identity") +
  ylab("Fraction of cells as putative doublets") +
  geom_hline(yintercept = 0.30, color="red", lty="dashed") +
  theme_classic() +
  theme(axis.text.x = element_text(angle=45, hjust = 0.9))

p

spurious_clust <- as.character(doub_enrich$Clusters_qc[doub_enrich$Doub_freq > 0.3])

spurious_cells <- proj$cellNames[proj$Clusters_qc %in% spurious_clust]

length(spurious_cells[!(spurious_cells %in% put_doublets)])

write(spurious_clust, "002_doublets/Mou_Cer_clustersEnriched_in_putDoublets.txt")
write(spurious_cells, "002_doublets/Mou_Cer_cells_in_clustersEnriched_in_putDoublets.txt")
```


```{r}
ggplot(filter(barcodes_init , Doublet_1stPass==F), aes(Sample, fill=Doublet_2ndPass)) +
  geom_bar() +
  coord_flip()
```

Let's save the project

```{r}
saveArchRProject(proj)
```

Removing doublets to create the final project

```{r}
subsetArchRProject(proj,
                   cells = proj$cellNames[!(proj$cellNames %in% c(put_doublets, spurious_cells))],
                   dropCells = F,
                   outputDirectory = "proj2_finalCells")

barcodes_init$PassQC_all <- barcodes_init$barcode %in% proj$cellNames
```

```{r}
proj <- loadArchRProject("proj2_finalCells/")
```

Now we can reprocess the project (Iterative LSI, harmony, clustering and UMAP):

```{r}
proj <- addIterativeLSI(ArchRProj = proj,
                        useMatrix = "TileMatrix",
                        name = "IterativeLSI_finalNucSet",
                        iterations=5,
                        clusterParams = list(
                          resolution = c(0.1, 0.2, 0.4, 0.8), 
                          sampleCells = 20000, 
                          n.start = 10
                        ), 
                        varFeatures = 100000,
                        dimsToUse = 1:100,
                        totalFeatures = 500000,
                        seed = 1,
                        LSIMethod = 1,
                        scaleDims = FALSE,
                        corCutOff = 0.75,
                        excludeChr = c("chrX", "chrY", "chrM"),
                        binarize = T,
                        force = T)
```

```{r}
proj <- addHarmony(proj,
                   reducedDims = "IterativeLSI_finalNucSet",
                   name ="HarmonyLSI_finalNucSet")
```

```{r}
proj <- addClusters(input = proj,
                    name = "Clusters_finalNucSet_res1.0",
                    reducedDims = "HarmonyLSI_finalNucSet",
                    method = "Seurat",
                    force = T,
                    resolution=1.0,
                    corCutOff = 0.75,
                    scaleDims = FALSE,
                    seed = 1)
```

We can also run a UMAP:

```{r}
proj <- addUMAP(ArchRProj = proj,
                name = "UMAP_finalNucSet",
                reducedDims = "HarmonyLSI_finalNucSet",
                minDist = 0.2,
                metric = "cosine",
                nNeighbors = 25,
                force = T,
                seed = 1,
                scaleDims = F,
                corCutOff = 0.75)
```

```{r, fig.width=8, fig.height=8}
p1 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Sample_short", embedding = "UMAP_finalNucSet", pal = colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample_short))), rastr = T, labelMeans = T)
p2 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Clusters_finalNucSet_res1.0", embedding = "UMAP_finalNucSet", rastr = T)

ggAlignPlots(p1, p2, type = "h")
```

```{r, fig.width=10, fig.height=8}
p1
```

Let's check again some interesting genes:

```{r}
p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name = interesting_genes, 
    embedding = "UMAP_finalNucSet",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)

p
```

```{r}
## Doublet score
p <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "DoubletScore", embedding = "UMAP_finalNucSet", rastr = T)
p


## Doublet enrichment
p <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "DoubletEnrichment", embedding = "UMAP_finalNucSet", rastr = T)
p

## nFragments
p <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "nFrags", embedding = "UMAP_finalNucSet", rastr = T)
p
```

We should save the LSI, UMAP and Cluster information:

```{r}
if (dir.exists("003_finalNucSet")==F){
  dir.create("003_finalNucSet")
}

## UMAP
saveRDS(getEmbedding(proj, embedding = "UMAP_finalNucSet", returnDF = F), "003_finalNucSet/Opossum_Cereb_finalNucSet_UMAP_merged_tiles.rds")

## LSI
saveRDS(getReducedDims(proj, reducedDims = "IterativeLSI_finalNucSet", returnMatrix = F), "003_finalNucSet/Opossum_Cereb_finalNucSet_LSI_merged_tiles.rds")

## Clusters and other cell data
clust <- getCellColData(proj) %>%
  as.data.frame() %>%
  rownames_to_column("cellName")

saveRDS(clust, "003_finalNucSet/Opossum_Cereb_finalNucSet_Clusters_merged_tiles_res1.0_and_barcodeStats.rds")
```

Let's save the project for now:

```{r}
saveArchRProject(proj)
```

Some QC plots:

TSS ridge plot

```{r}
p <- plotGroups(
    ArchRProj = proj, 
    groupBy = "Sample_short", 
    colorBy = "cellColData", 
    name = "TSSEnrichment",
    pal = rev(colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample_short)))),
    plotAs = "ridges"
   ) +
  ylab("Samples")

p

pdf("Figures/001_QC/TSSenrichment_byLib.pdf", width = 6, height = 6, useDingbats = F); print(p);dev.off()
```

Fragment size distribution

```{r}
fragment_sizes <- plotFragmentSizes(ArchRProj = proj, returnDF = T)

head(fragment_sizes)
dim(fragment_sizes)

fragment_sizes <- as.data.frame(fragment_sizes) %>%
  dplyr::rename(Sample=sampleName) %>%
  left_join(unique(dplyr::select(as.data.frame(getCellColData(proj)), Sample, Sample_short)))

p <-ggplot(fragment_sizes, aes(x=fragmentSize, y=fragmentPercent, color=as.factor(Sample_short))) +
  geom_line() +
  theme_classic() +
  xlab("Fragment size") +
  ylab("Density") +
  theme(legend.position = "bottom")+
  scale_color_manual(values = colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample_short))), name="Sample")

p

pdf("Figures/001_QC/FragLength_distribution_byLib.pdf", width = 6, height = 6, useDingbats = F); print(p);dev.off()

```

TSS enrichment

```{r}
tssEnr <- plotTSSEnrichment(ArchRProj = proj, returnDF = T)

head(tssEnr)
dim(tssEnr)

tssEnr <- as.data.frame(tssEnr) %>%
  dplyr::rename(Sample=sampleName) %>%
  left_join(unique(dplyr::select(as.data.frame(getCellColData(proj)), Sample, Sample_short)))

p <-ggplot(tssEnr, aes(x=x, y=smoothValue, color=as.factor(Sample_short))) +
  geom_line() +
  theme_classic() +
  xlab("Distance from TSS (bp)") +
  ylab("Normalised signal") +
  theme(legend.position = "bottom")+
  scale_color_manual(values = colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample_short))), name="Sample")

p

pdf("Figures/001_QC/TSS_signal_distribution_byLib.pdf", width = 6, height = 6, useDingbats = F); print(p);dev.off()
```

Now, we want to assign cell types to our clusters. For that we will need to identify marker genes.

```{r}
maj_clust <- names(which(table(proj$Clusters_finalNucSet_res1.0) >= 30))

maj_clust

markersGS_clusters_broad_cell_types <-getMarkerFeatures(proj,
                  groupBy = "Clusters_finalNucSet_res1.0",
                  useMatrix = "GeneScoreMatrix",
                  useGroups = maj_clust,
                  threads = 10,
                  scaleTo = 10^4)

saveRDS(markersGS_clusters_broad_cell_types, "003_finalNucSet/Opossum_Cereb_finalNucSet_Clusters_merged_tiles_res1.0_markerGenes.rds")
```

Exporting the barcode stats.

```{r}
saveRDS(barcodes_init, "001_barcode_qc/Opossum_Cerebellum_barcode_stats.rds")
```


```{r}
sessionInfo()
```