---
title: "Opossum Cerebellum 001 - Barcode QC with ArchR"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ArchR)
  library(tidyverse)
  library(RColorBrewer)
})
```

```{r}
addArchRThreads(threads = 20) 

## Working directory. This is where the arrow files will be generated and stored.
setwd("~/Opo_Cereb/")

proj <- loadArchRProject("~/Opo_Cereb/proj2_finalCells/")

markersGS_clusters_broad_cell_types <- readRDS("003_finalNucSet/Human_Cereb_finalNucSet_Clusters_merged_tiles_res1.0_markerGenes.rds")

markersGS_cellAnno <- getMarkers(markersGS_clusters_broad_cell_types, cutOff = "FDR < 0.05 & Log2FC > 1.25")
```

Exploring clusters and known marker genes

```{r, fig.width=8, fig.height=8}
p1 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Sample_short", embedding = "UMAP_finalNucSet", pal = rev(colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample_short)))), rastr = T, labelMeans = F)
p2 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Clusters_finalNucSet_res1.0", embedding = "UMAP_finalNucSet", rastr = T)

ggAlignPlots(p1, p2, type = "h")
```

```{r}
interesting_genes <- toupper(c("Atoh1", "Sox2", "Notch1","Cyp26b1","Foxp2", "Skor2", "Car8", "Cdon", "Gad2", "Sox14", "Tlx3", "Lmx1a", "Eomes", "Mgat5","Pax2", "Top2a", "Ptf1a", "Kirrel2", "Rora", "Rorb", "Slc17a6", "Meis2", "Aqp4", "Olig2", "Cbln3", "Fgf3", "Grin2b", "Neurod1", "Nhlh1", "Gdf10", "Fabp7", "Map3k1", "Gli2", "Esrrg", "Sorcs3", "Cdh22", "Etv1", "Junb", "Mef2a", "Foxp1", "Cdh9", "March11", "Pvalb", "Lhx2", "Pax5", "Zfhx3", "Zfhx4", "Fgf17", "Syndig1l","Gabra6", "Slc1a3", "Inpp5d", "Cx3cr1", "Mobp", "Sox10", "Lims2", "Slc4a1", "Angptl8", "Lmx1b", "Isl1", "Lef1"))

interesting_genes <- interesting_genes[interesting_genes %in% proj@geneAnnotation$genes$symbol]

p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneScoreMatrix", 
    name = interesting_genes, 
    embedding = "UMAP_finalNucSet",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)

p
```

```{r}
inspect_clust <- function(clust,
                          archrproj=proj,
                          slot="Clusters_finalNucSet_res1.0",
                          markers=markersGS_cellAnno,
                          umap="UMAP_finalNucSet") {
  
  ## Index of cells belonging to the cluster
  i <- which(getCellColData(archrproj, slot,drop = T)==clust)
  print("Total cells")
  print(length(i))
  print("--------------------")
  ## Timepoint distribution
  print("Timepoint distribution")
  print(sort(table(archrproj$Timepoint[i]), decreasing = T))
  print(sort(table(archrproj$Sample_short[i]), decreasing = T))
  
  if (length(i) >= 30) {
     ## Gettting top 20 markers, excluding LncRNAs and miRNAs with non-informative names
      m <- markers[[clust]]$name[grepl("^Gm[0-9]|Rik$|^Mir[0-9]|XLOC|ENSMODG", markers[[clust]]$name)==F]
  
      if(length(m > 20)) {
        m <- m[1:20]
      }
  
      print("Top 20 markers")
      print(m)
      print("--------------------")
  
      ## Plotting top genes
      print("Plotting top marker genes")
      p <- plotEmbedding(
        ArchRProj = archrproj, 
        colorBy = "GeneScoreMatrix", 
        name =m[is.na(m)==F], 
        embedding = umap,
        quantCut = c(0.01, 0.95),
        imputeWeights = NULL,
      )
      print(p)
      print("--------------------")
      } else {
        print("Too few cells to get markers")
        }
}
```

Initiating a "Cell_type" variable for our cells

```{r}
proj$Cell_type <- NA
```


```{r}
inspect_clust("C1")
```

Too few cells, keeping them as "Other"

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C1"] <- "Other"
```

```{r}
inspect_clust("C2")
```

C2: Astroglia (mixture of astroblasts and adult astrocytes)

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C2"] <- "Astroglia"
```

```{r}
inspect_clust("C3")
```

C3: Astrocytes (WM) (part of astroglia)

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C3"] <- "astro_parenhymal"
```

```{r}
inspect_clust("C4")
```

C4: Microglia

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C4"] <- "Microglia"
```

```{r}
inspect_clust("C5")
```

```{r}
oligo_m <- toupper(c("Mobp", "Sox10", "Olig2", "Lims2", "Gpr17", "Cspg4", "Mag"))
  
oligo_m <- oligo_m[oligo_m %in% proj@geneAnnotation$genes$symbol]

plotEmbedding(
        ArchRProj = proj, 
        colorBy = "GeneScoreMatrix", 
        name =oligo_m, 
        embedding = "UMAP_finalNucSet",
        quantCut = c(0.01, 0.95),
        imputeWeights = NULL,
      )
```
C5: OPC 

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C5"] <- "OPC"
```


```{r}
inspect_clust("C6")
```

C6: OPC 

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C6"] <- "OPC"
```

```{r}
inspect_clust("C7")
```

C7: oligodendrocytes

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C7"] <- "Oligodendrocyte"
```

```{r}
inspect_clust("C8")
```

C8: Vascular

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C8"] <- "Vascular"
```

```{r}
inspect_clust("C9")
```

C9: Purkinje cells

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C9"] <- "PC"
```

```{r}
inspect_clust("C10")
```

C10: Interneuron (GL)

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C10"] <- "Interneuron_late_GL"
```


```{r}
inspect_clust("C11")
```

C11: IN (ML/PL) aka "late"

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C11"] <- "Interneuron_late_ML/PL"
```

```{r}
inspect_clust("C12")
```


```{r}
gc_m <- toupper(c("Pax6", "Atoh1", "Barhl1", "Gli2", "Nhlh1","Neurod1","Tmem71", "Cog7", "Pou2af1","Egfr","Etv1", "Cbln3", "Nfix", "Grin2b", "Gabra6", "Cadps2"))
  
gc_m <- gc_m[gc_m %in% proj@geneAnnotation$genes$symbol]

plotEmbedding(
        ArchRProj = proj, 
        colorBy = "GeneScoreMatrix", 
        name =gc_m, 
        embedding = "UMAP_finalNucSet",
        quantCut = c(0.01, 0.95),
        imputeWeights = NULL,
      )
```

```{r}
inspect_clust("C13")
```

C12-C13: GNPs

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0 %in% c("C12", "C13")] <- "GCP"
```

```{r}
inspect_clust("C14")
```

C14: IN

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0=="C14"] <- "Interneuron_P21"
```

```{r}
inspect_clust("C15")
```

C15: migrating GCs

```{r}
inspect_clust("C16")
```

C16: IGL (mature GCs in P21)

However, both C15 and C16 seem to contain traces of UBCs. Let's see if we can pick them out by subclustering.

```{r}
c15c16 <- subsetCells(proj, cellNames = proj$cellNames[proj$Clusters_finalNucSet_res1.0 %in% c("C15", "C16")])

c15c16 <- addClusters(input = c15c16,
                    name = "Subclusters",
                    reducedDims = "HarmonyLSI_finalNucSet",
                    method = "Seurat",
                    force = T,
                    resolution=0.6,
                    corCutOff = 0.75,
                    scaleDims = FALSE,
                    seed = 1)

c15c16  <- addUMAP(ArchRProj = c15c16,
                name = "UMAP_sub",
                reducedDims = "HarmonyLSI_finalNucSet",
                minDist = 0.2,
                metric = "cosine",
                nNeighbors = 25,
                force = T,
                seed = 1,
                scaleDims = F,
                corCutOff = 0.75)

plotEmbedding(ArchRProj = c15c16, colorBy = "cellColData", name = "Subclusters", embedding = "UMAP_sub", rastr = T)
```

```{r}
gcVsUBC <- toupper(c("Lmx1a", "Trpc3", "Etv1", "Cbln3", "Pax6", "Cadps2", "Gabra6", "Pax6"))

gcVsUBC <- gcVsUBC[gcVsUBC %in% c15c16@geneAnnotation$genes$symbol]

p <- plotEmbedding(
    ArchRProj = c15c16, 
    colorBy = "GeneScoreMatrix", 
    name = gcVsUBC, 
    embedding = "UMAP_sub",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)

p

sum(c15c16$Subclusters=="C3")
table(c15c16$Sample_short[c15c16$Subclusters=="C3"])
table(c15c16$Clusters_finalNucSet_res1.0[c15c16$Subclusters=="C3"])
```

Cluster 3 here is Lmx1a+ -> UBC

```{r}
ubc_names <- c15c16$cellNames[c15c16$Subclusters=="C3"] 

proj$Cell_type[proj$Clusters_finalNucSet_res1.0 %in% c("C15", "C16") & !(proj$cellNames %in% ubc_names)] <- "GC_diff"

proj$Cell_type[proj$Clusters_finalNucSet_res1.0 %in% c("C15", "C16") & proj$cellNames %in% ubc_names] <- "UBC"
```

```{r}
inspect_clust("C17")
```

```{r}
inspect_clust("C18")
```

C18-22: no real markers but show activity for known GC markers:

```{r}
plotEmbedding(
        ArchRProj = proj, 
        colorBy = "GeneScoreMatrix", 
        name =gc_m, 
        embedding = "UMAP_finalNucSet",
        quantCut = c(0.01, 0.95),
        imputeWeights = NULL,
      )
```

```{r}
proj$Cell_type[proj$Clusters_finalNucSet_res1.0 %in% c("C17", "C18", "C19", "C20", "C21", "C22")] <- "GC_mature"
```

```{r}
sum(is.na(proj$Cell_type))
sort(table(proj$Cell_type), decreasing = T)

unique(proj$Cell_type)

cell_hierarchy <- data.frame(Cell_type=unique(proj$Cell_type),
           Cell_type_broad=c("Interneuron", "GC", "PC", "Astroglia", "GC", "GC", "UBC", "Other", "Oligodendrocyte", "Microglia", "Oligodendrocyte", "Other", "Interneuron", "Interneuron", "Astroglia"), stringsAsFactors = F)

cell.info <- getCellColData(proj) %>%
  as.data.frame() %>%
  left_join(cell_hierarchy)

sum(cell.info$Cell_type==proj$Cell_type)

proj$Cell_type_broad <- cell.info$Cell_type_broad
```

```{r, fig.width=8, fig.height=8}
p1 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Sample_short", embedding = "UMAP_finalNucSet", pal = rev(colorRampPalette(brewer.pal(11, "Spectral"))(length(unique(proj$Sample_short)))), rastr = T, labelMeans = F)
p2 <- plotEmbedding(ArchRProj = proj, colorBy = "cellColData", name = "Cell_type", embedding = "UMAP_finalNucSet", rastr = T)

ggAlignPlots(p1, p2, type = "h")
```

```{r}
saveArchRProject(proj)
```

```{r}
sessionInfo()
```

