---
title: "Mouse to opossum CRE specificity"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(RColorBrewer)
})
```

Here we want to investigate whether conserved CREs retain the same cell type-specificities:

```{r}
opo_pseudo <- read.table("~/Opo_Cereb/004_peaks/Opossum_lineage_by_timepoint_pseudobulk_cpm_robust.txt", header = T, sep = "\t", stringsAsFactors = F)

mouse_pseudo <- read.table("~/Mouse_Cereb/013_globalCREactivity/Mouse_lineage_by_timepoint_pseudobulk_cpm.txt", header = T, sep = "\t", stringsAsFactors = F)

## Selecting only relevant timepoints in mouse (mouse P4 ~  opo P21  and mouse P63 ~ opo adult)
mouse_pseudo <- mouse_pseudo[, grepl("_P04|_P63", colnames(mouse_pseudo))]

## Loading CRE orthology table
ortho_cre <- read_tsv("~/Opo_Cereb/004_peaks/Opossum-Mouse.reciprocal.bestMatch.txt")
## Limiting to only robust peaks in both species
nrow(ortho_cre)
ortho_cre_rob <- filter(ortho_cre, 
                        Mouse %in% row.names(mouse_pseudo),
                        Opossum %in% row.names(opo_pseudo))

nrow(ortho_cre_rob)

peak2gene <- readRDS("~/Mouse_Cereb/Revisions/002_cre2gene_assignment/CRE_to_gene_assignment_sigInteractions.rds")
```

Subsetting to shared samples

```{r}
## Identifying samples in opossum
opo_samples <- colnames(opo_pseudo)
opo_samples

opo_samples <- c("Astroglia_P21", "Astroglia_adult", "PC_P21", "Interneuron_P21", "Interneuron_adult", "GC_P21", "GC_adult", "UBC_P21", "Oligodendrocyte_adult", "Microglia_adult")
opo_pseudo <- opo_pseudo[,opo_samples]
## Converting to mouse names
mou_samples <- gsub("adult", "P63", opo_samples)
mou_samples <- gsub("P21", "P04", mou_samples)

mouse_pseudo <- mouse_pseudo[, mou_samples]
```


Now we can limit our pseudobulks to the orthologous CREs:

```{r, fig.width=6, fig.height=5}
opo_pseudo.cons <- opo_pseudo[ortho_cre_rob$Opossum,]
mouse_pseudo.cons <- mouse_pseudo[ortho_cre_rob$Mouse,]

s <- cor(mouse_pseudo.cons, opo_pseudo.cons, method = "spearman")

pheatmap::pheatmap(s, cluster_rows = F, cluster_cols = F)
```

Let's also subset to distal CREs only:

```{r, fig.width=6, fig.height=5}
mouse_peaks <- read_tsv("~/Mouse_Cereb/006_peaks/Mouse_Cerebellum_reproducible_peaks_summary.txt")
mouse_peaks <- filter(mouse_peaks, robust==T)

opo_peaks <- read_tsv("~/Opo_Cereb/004_peaks/Opossum_Cerebellum_reproducible_peaks_summary.txt")
opo_peaks <- filter(opo_peaks, robust==T)

ortho_cre_rob_dist <- filter(ortho_cre_rob,
                             Mouse %in% mouse_peaks$peak[mouse_peaks$genomic_class_broad %in% c("Distal") & mouse_peaks$robust==T],
                             Opossum %in% opo_peaks$peak[opo_peaks$genomic_class_broad %in% c("Distal") & opo_peaks$robust==T])

opo_pseudo.cons.dist <- opo_pseudo[ortho_cre_rob_dist$Opossum,]
mouse_pseudo.cons.dist <- mouse_pseudo[ortho_cre_rob_dist$Mouse,]

s.dist <- cor(mouse_pseudo.cons.dist, opo_pseudo.cons.dist, method = "spearman")

pheatmap::pheatmap(s.dist, cluster_rows = F, cluster_cols = F)

draw(Heatmap(matrix = t(s.dist),
        cluster_rows =F,
        cluster_columns = F,
        show_row_names = T,
        border = TRUE,
        col = colorRampPalette(rev(brewer.pal(n=9,name = "RdBu")))(20),
        heatmap_legend_param = list( title = "Spearman's rho")), heatmap_legend_side = "left", annotation_legend_side = "right")
```

```{r}
pdf("~/Mouse_Cereb/Revisions/Figures/009_Mouse_opossum_intergenicCREonly_Spearman_rho.pdf", width=6, height = 5, useDingbats = F);draw(Heatmap(matrix = t(s.dist),
        cluster_rows =F,
        cluster_columns = F,
        show_row_names = T,
        border = TRUE,
        col = colorRampPalette(rev(brewer.pal(n=9,name = "RdBu")))(20),
        heatmap_legend_param = list( title = "Spearman's rho")), heatmap_legend_side = "left", annotation_legend_side = "right") ; dev.off()
```


```{r, fig.width=4, fig.height=3}
c <- diag(cor(t(opo_pseudo[ortho_cre_rob_dist$Opossum,]), t(mouse_pseudo[ortho_cre_rob_dist$Mouse,])))

c_shuffled <- diag(cor(t(mouse_pseudo[ortho_cre_rob_dist$Mouse,]), t(opo_pseudo[sample(ortho_cre_rob_dist$Opossum, length(ortho_cre_rob_dist$Opossum), replace = F),])))

hist(c)
hist(c_shuffled)

df <- data.frame(cor=c(c, c_shuffled),
                 type=c(rep("ortholog", length(c)), rep("shuffled", length(c_shuffled))),
                 stringsAsFactors = F)


p <- ggplot(df, aes(cor, fill=type, color=type)) +
  geom_density(alpha=0.2, lwd=0.8) +
  scale_fill_manual(values = c("deepskyblue3", "darkorange")) +
  scale_color_manual(values = c("deepskyblue3", "darkorange")) +
  xlab("Pearson's correlation") +
  theme_classic()

p

pdf("~/Mouse_Cereb/Revisions/Figures/009_Mouse_opossum_intergenicCREonly_Pearsons_acrossStages.pdf", width=4, height = 3, useDingbats = F);print(p); dev.off()
```

Exporting the correlations

```{r}
ortho_cre_rob_dist$correlation <- c

write_tsv(ortho_cre_rob_dist, "~/Mouse_Cereb/Revisions/009_opossum_comparisons/Mouse_opossum_orthoCREs_intergenicInBoth_withCorr.txt")
```

Also giving general estimates of conservation.

Here it's important to only consider mouse CREs that are active in stages and cell types with correspondence to our targeted opossum dataset.

We can use our pseudobulks and ask for mouse CREs reaching at least 5 CPM in at least one sample (with a corresponding opossum sample):

```{r}
mouse_peaks_expr <- filter(mouse_peaks, peak %in% row.names(mouse_pseudo)[rowSums(mouse_pseudo >= 5) > 0])

mou2opo <- read_tsv("~/Opo_Cereb/004_peaks/mouse.peaks.overTo.opossum.bed", col_names = F)

mouse_peaks_expr$conservation <- ifelse(mouse_peaks_expr$peak %in% ortho_cre$Mouse, yes = "conserved-activity", no = ifelse(mouse_peaks_expr$peak %in% mou2opo$X4, yes = "conserved-sequence", no = "mouse-specific"))
table(mouse_peaks_expr$conservation)

## All robust mouse peaks
table(mouse_peaks$genomic_class_broad)/nrow(mouse_peaks)
## Mouse peaks expressed in corresponding stages in opossum
table(mouse_peaks_expr$genomic_class_broad)/nrow(mouse_peaks_expr)
## Mouse peaks expressed in corresponding stages and with conserved sequence in opossum
table(mouse_peaks_expr$genomic_class_broad[mouse_peaks_expr$conservation!="mouse-specific"])/nrow(mouse_peaks_expr[mouse_peaks_expr$conservation!="mouse-specific",])
## Mouse peaks expressed in corresponding stages and with conserved activity in opossum
table(mouse_peaks_expr$genomic_class_broad[mouse_peaks_expr$conservation=="conserved-activity"])/nrow(mouse_peaks_expr[mouse_peaks_expr$conservation=="conserved-activity",])

mouse_stats <- group_by(mouse_peaks, genomic_class_broad) %>%
  summarise(
    all=n(),
    expr=sum(peak %in% mouse_peaks_expr$peak),
    seq_cons=sum(peak %in% mouse_peaks_expr$peak[mouse_peaks_expr$conservation!="mouse-specific"]),
    act_cons=sum(peak %in% mouse_peaks_expr$peak[mouse_peaks_expr$conservation=="conserved-activity"])) %>% gather(-genomic_class_broad, key = "type", value = "n") %>%
  mutate(type=factor(type, levels = rev(c("all", "expr", "seq_cons", "act_cons"))),
         genomic_class_broad=factor(genomic_class_broad, levels = rev(c("Promoter", "Exonic", "Intronic", "Distal"))))

mouse_stats <- group_by(mouse_stats, type) %>%
  mutate(f=round(n/sum(n), digits = 3) * 100) %>%
  ungroup()

ggplot(mouse_stats, aes(x=type, y=n, fill=genomic_class_broad)) +
  geom_col() +
  geom_text(aes(label=paste0(f, "%")), position = position_stack(vjust = 0.5)) +
  scale_fill_manual(values = c("darkorange2",  "darkolivegreen3","deepskyblue3","darkorchid"))+
  coord_flip() +
  theme_classic()

p1 <- ggplot(mouse_stats, aes(x=type, y=n, fill=genomic_class_broad)) +
  geom_col(position = "fill") +
  geom_text(aes(label=n), position = position_fill(vjust = 0.5)) +
  scale_fill_manual(values = c("darkorange2",  "darkolivegreen3","deepskyblue3","darkorchid"))+
  ylab("Fraction") +
  coord_flip() +
  theme_classic()

p1

mouse_stats_agr <- group_by(mouse_stats, type) %>%
  summarise(n=sum(n))

p2 <- ggplot(mouse_stats_agr, aes(x=type, y=n)) +
  geom_col(fill="gray70") +
  geom_text(aes(label=n), position = position_identity()) +
  ylab("Count") +
  coord_flip() +
  theme_classic()

p2

gridExtra::grid.arrange(p1, p2, nrow=1, widths=c(2, 1))


pdf("~/Mouse_Cereb/Revisions/Figures/009_Mouse_Peaks_conservedInOpossum.pdf", width=8, height = 3, useDingbats = F);gridExtra::grid.arrange(p1, p2, nrow=1, widths=c(2, 1)); dev.off()

ggplot(filter(mouse_peaks_expr, genomic_class=="Distal"), aes(x=conservation, y=meanPhastCons_100bp)) +
  geom_violin() +
  geom_boxplot(width=0.05, notch = T, outlier.alpha = 0) +
  theme_classic()
```

```{r}
sessionInfo()
```

