---
title: "Mouse Cerebellum - Adult cerebellum: Repeat classes"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

Here we try to associate the younger age and lower constraint of microglia CREs with specific repeat classes.

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(RColorBrewer)
  library(viridis)
  library(SummarizedExperiment)
})
```

Loading our peak matrix and info summary

```{r}
peak_mat <- readRDS("~/Mouse_Cereb/006_peaks/Mouse_Cerebellum_reproducible_peak_matrix_SE_withInfo_robustPeaks.rds")

peak.info <- read.table("~/Mouse_Cereb/006_peaks/Mouse_Cerebellum_reproducible_peaks_summary.txt", header = T, sep = "\t", stringsAsFactors = F)
```

Repeats by class

```{r}
rep_class <- na.omit(unlist(sapply(peak.info$overlappingRepClass, function(x) str_split(x, ",", simplify = T))))
rep_family <- na.omit(unlist(sapply(peak.info$overlappingRepFamily, function(x) str_split(x, ",", simplify = T))))

sort(table(rep_class), decreasing = T)
sort(table(rep_family), decreasing = T)[1:10]

major_reps <- names(sort(table(rep_family), decreasing = T)[!(names(sort(table(rep_family), decreasing = T)) %in% c("Low_complexity", "Simple_repeat"))][1:10])

major_reps
```

Fraction of reads in peaks from each repeat class per cell:

```{r}
cell_i <- which(peak_mat@colData$Timepoint=="P63")

atac.data <- assay(peak_mat[,cell_i])
row.names(atac.data) <- peak_mat@elementMetadata$peak

rep_counts <- do.call(rbind, lapply(major_reps, function(r) {
  peak_i <- peak.info$peak[grepl(paste0(r, "$|", r, ","), peak.info$overlappingRepFamily)]
  peak_i <- peak_i[peak_i %in% row.names(atac.data)]
  return(Matrix::colSums(atac.data[peak_i,]))
}))

rep_freq <- t(t(rep_counts)/Matrix::colSums(atac.data))

summary(colSums(rep_freq))

row.names(rep_freq) <- major_reps

```

Adding the cell type annotation

```{r}
cell_anno <- readRDS("~/Mouse_Cereb/004_cellTypes_broad/Mouse_Cerebellum_atac_finalCellType_annotation.rds")

rep_freq.gather <- as.data.frame(rep_freq) %>%
  rownames_to_column("repFamily") %>%
  gather(key = "cell", value = "freq", -repFamily) %>%
  left_join(cell_anno)
```

```{r, fig.width=12, fig.height=4}
repeat.info <- read_tsv("~/Mouse_Cereb/006_peaks/001_peak_annotation/repeatmasker/mm10_Repeatmasker_info.txt")

table(repeat.info$repName[repeat.info$repFamily=="Alu"])

rep_freq.gather <-select(repeat.info, repFamily, repClass) %>%
  unique() %>%
  right_join(rep_freq.gather)

rep_freq.gather$repFamily[rep_freq.gather$repFamily=="Alu"] <- "B1"

rep_freq.gather <- arrange(rep_freq.gather, repClass, repFamily) %>%
  mutate(repClass=factor(repClass, levels = unique(repClass)),
         repFamily=factor(repFamily, levels = unique(repFamily)))

p <- ggplot(filter(rep_freq.gather,Cell_type_broad %in% c("Astroglia", "GC", "Interneuron", "Microglia", "Oligodendrocyte")), aes(x=repFamily, y=freq, fill=Cell_type_broad)) +
  geom_boxplot(outlier.alpha = 0) +
  scale_fill_manual(values = c("#BB4D90", "#62943c", "#d3405a", "#44be8d", "#A692E8")) +
  theme_classic() +
  ylab("Fraction of counts in peaks overlapping repeat") +
  xlab("")

p

pdf("~/Mouse_Cereb/Figures/008_conservation/TE_byCellType_P63_microglia.pdf", width = 12, height = 4, useDingbats = F); print(p); dev.off()
```

```{r}
sort(sapply(levels(rep_freq.gather$repFamily), function(r) {
  return(wilcox.test(rep_freq.gather$freq[rep_freq.gather$Cell_type_broad=="Microglia" & rep_freq.gather$repFamily==r],
            rep_freq.gather$freq[rep_freq.gather$Cell_type_broad %in% c("Astroglia", "GC", "Interneuron", "Oligodendrocyte") & rep_freq.gather$repFamily==r], alternative = "greater")$p.value)
}))

sort(sapply(levels(rep_freq.gather$repFamily), function(r) {
  mean(rep_freq.gather$freq[rep_freq.gather$Cell_type_broad=="Microglia" & rep_freq.gather$repFamily==r])/
            mean(rep_freq.gather$freq[rep_freq.gather$Cell_type_broad %in% c("Astroglia", "GC", "Interneuron", "Oligodendrocyte") & rep_freq.gather$repFamily==r])
}), decreasing = T)
```

We see that microglia are enriched for recently expanded TE classes in the mouse genome, such as B1, B2, ERV elements and L1 but not for more ancient expansions, such as MIRs, L2s and DNA repeats.

```{r}
saveRDS(rep_freq.gather, "~/Mouse_Cereb/016_conservation_and_repeats/Mouse_Cerebellum_P63_repeatClasses_byCell.rds")
```


```{r}
sessionInfo()
```

