---
title: "FANTOM5 integration"
author: "Ioannis Sarropoulos"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(GenomicRanges)
  library(SummarizedExperiment)
})
```

Here we want to integrate with the data from the FANTOM5 project.

##1. Overlap with cerebellum peaks

The expectation here is that we preferentially capture eRNAs active in the cerebellum.

```{r}
## Our ATAC peaks
peak.info <- read_tsv("~/Mouse_Cereb/006_peaks/Mouse_Cerebellum_reproducible_peaks_summary.txt")

peak.gr <- GRanges(seqnames = peak.info$chr,
              ranges = IRanges(
                start = peak.info$start,
                end = peak.info$end))

peak.gr$peak <- peak.info$peak
```

```{r}
## eRNA coordinates
eRNA.bed <- read_tsv("~/Resources/eRNA_Fantom5/F5.mm10.enhancers.bed", col_names = F)

eRNA.gr <- GRanges(seqnames = eRNA.bed$X1,
              ranges = IRanges(
                start = eRNA.bed$X2,
                end = eRNA.bed$X3))

eRNA.gr$eRNA <- eRNA.bed$X4
```

Overlapping the two

```{r}
eRNA_overlaps <- findOverlaps(peak.gr, eRNA.gr)

## Using all peaks
length(eRNA_overlaps)
length(unique(queryHits(eRNA_overlaps)))
length(unique(subjectHits(eRNA_overlaps)))

length(unique(subjectHits(eRNA_overlaps)))/length(eRNA.gr)
length(unique(subjectHits(eRNA_overlaps)))/length(eRNA.gr)

## Using robust peaks
robust_i <- which(peak.gr$peak %in% peak.info$peak[peak.info$robust])
length(unique(subjectHits(eRNA_overlaps[queryHits(eRNA_overlaps) %in% robust_i])))
length(unique(subjectHits(eRNA_overlaps[queryHits(eRNA_overlaps) %in% robust_i])))/length(eRNA.gr)
```

We capture approximately 50% of their eRNAs (37% in robust peaks):

The expectation is that these eRNAs will be more active in the cerebellum than in other tissues:

```{r}
mcols(eRNA.gr)$class <- "notFound"
mcols(eRNA.gr[unique(subjectHits(eRNA_overlaps))])$class <- "permissive"
mcols(eRNA.gr[unique(subjectHits(eRNA_overlaps[queryHits(eRNA_overlaps) %in% robust_i]))])$class <- "robust"           

table(mcols(eRNA.gr)$class)
```

Activity matrix for eRNAs (we will use the binarised matrix that accounts for library noise).

```{r}
eRNA.mat <- read.table("~/Resources/eRNA_Fantom5/F5.mm10.enhancers.expression.usage.matrix", header = T, sep = "\t", stringsAsFactors = F)

dim(eRNA.mat)

colnames(eRNA.mat)

summary(colSums(eRNA.mat))

eRNA.sample.info <- read_tsv("~/Resources/eRNA_Fantom5/Mouse.sample_name2library_id.txt", col_names = c("description", "sample"))

## Identifying samples associated with cerebellum
eRNA.sample.info$class <- ifelse(grepl("cerebel|Granule cells", eRNA.sample.info$description),
                                 yes = "cerebellum",
                                 no = ifelse(grepl("brain|neur|astrocyte|microglia|medulla oblongata|hippocampus|pituitary|encephalon|spinal cord|eye|cortex|striatum", eRNA.sample.info$description, ignore.case = T) & grepl("embryonic stem cells", eRNA.sample.info$description)==F, 
                                             yes ="other neural", 
                                             no = ifelse(grepl("whole body, embryo", eRNA.sample.info$description),
                                                         yes = "whole embryo" , no = "other")))

eRNA.sample.info$class <- factor(eRNA.sample.info$class, levels = c("cerebellum", "other neural", "whole embryo", "other"))
  
table(eRNA.sample.info$class)
```

```{r, fig.width=2.5, fig.height=4.5}
eRNA.stats <- data.frame(
  sample=colnames(eRNA.mat),
  totalEnhancers=colSums(eRNA.mat),
  enhancersInRobust=colSums(eRNA.mat[eRNA.gr$eRNA[eRNA.gr$class=="robust"],]),
  enhancersInPermissive=colSums(eRNA.mat[eRNA.gr$eRNA[eRNA.gr$class %in% c("robust", "permissive")],]),
  stringsAsFactors = F
) %>%
  mutate(fractionInRobust=enhancersInRobust/totalEnhancers,
         fractionInPermissive=enhancersInPermissive/totalEnhancers) %>%
  left_join(eRNA.sample.info)

summary(eRNA.stats$totalEnhancers)

p <- ggplot(filter(eRNA.stats, totalEnhancers >= 200), aes(x=class, y=fractionInRobust, fill=class)) +
  geom_boxplot(notch = T) + 
  theme_classic() +
  theme(axis.text.x = element_text(angle=45, hjust = 0.95)) +
  xlab("") +
  ylab("Fraction of active eRNAs in robust ATAC peaks") +
  scale_fill_manual(values = c("#33CCFF", "dodgerblue3", "deeppink3", "gray"), guide=F)

p

pdf("~/Mouse_Cereb/Revisions/Figures/003_FANTOM5_eRNAs_fractionInCerebellumCREs.pdf", width = 2.5, height = 4.5, useDingbats = F); print(p); dev.off()

## Stats
wilcox.test(eRNA.stats$fractionInRobust[eRNA.stats$class=="cerebellum"], eRNA.stats$fractionInRobust[eRNA.stats$class=="other neural"])

wilcox.test(eRNA.stats$fractionInRobust[eRNA.stats$class=="other neural"], eRNA.stats$fractionInRobust[eRNA.stats$class=="whole embryo"])

wilcox.test(eRNA.stats$fractionInRobust[eRNA.stats$class=="whole embryo"], eRNA.stats$fractionInRobust[eRNA.stats$class=="other"])

## Fraction of eRNAs captured in peaks
median(eRNA.stats$fractionInRobust[eRNA.stats$class=="cerebellum"])
median(eRNA.stats$fractionInRobust[eRNA.stats$class=="other"])
```

We clearly see that we are enriching for eRNAs active in the cerebellum!

##2. Correlation between eRNA and CAGE-TSS expression in cerebellum samples

This might be very tricky given the differences in sensitivity between libraries (very different sequencing depths and number of active enhancers), which means that there is a lot of technical zeroes here!

Let's look into the promoter data and see if the colSums are similar.

```{r}
summary(eRNA.stats$totalEnhancers[eRNA.stats$class=="cerebellum"])

cage_mat <- read.table("~/Resources/eRNA_Fantom5/mm10_fair+new_CAGE_peaks_phase1and2_counts_ann.osc.txt", header = T, sep = "\t", stringsAsFactors = F)


cage_anno <- cage_mat[, 1:7]
cage_mat <- cage_mat[, 8:ncol(cage_mat)]
row.names(cage_mat) <- cage_anno$X00Annotation
colnames(cage_mat)[1:10]

x <- str_extract(colnames(cage_mat), "CNhs[0-9]*")

length(x)
sum(x %in% eRNA.sample.info$sample)
x[!(x %in% eRNA.sample.info$sample)]

## Subsetting for samples were we have info
colnames(cage_mat) <- str_extract(colnames(cage_mat), "CNhs[0-9]*")
cage_mat <- cage_mat[, colnames(eRNA.mat)]

cor(log10(colSums(eRNA.mat)), log10(colSums(cage_mat)))

plot(log10(colSums(eRNA.mat)), log10(colSums(cage_mat)), pch=".")
```

Overlapping the CAGE peaks with our promoters

```{r}
prom.gr <- peak.gr[peak.gr$peak %in% peak.info$peak[peak.info$robust==T & peak.info$genomic_class=="protein_coding-Promoter"]]

cage_bed <- read_tsv("~/Resources/eRNA_Fantom5/mm10_fair+new_CAGE_peaks_phase1and2.bed", col_names = F)
cage.gr <- GRanges(seqnames = cage_bed$X1,
              ranges = IRanges(
                start = cage_bed$X2,
                end = cage_bed$X3))

mcols(cage.gr)$cage <- cage_bed$X4

cage_prom_overlaps <- findOverlaps(prom.gr, cage.gr)
length(cage_prom_overlaps)
length(unique(queryHits(cage_prom_overlaps)))
length(unique(subjectHits(cage_prom_overlaps)))
```
```{r}
dim(cage_mat)

sum(colnames(cage_mat) == colnames(eRNA.mat))

## Estimating scaling factor based on full CAGE data
sf <- edgeR::calcNormFactors(cage_mat, method="RLE")

summary(colSums(cage_mat))

cage_mat.tpm <- t(t(cage_mat) / colSums(cage_mat)) * 1e6
cage_mat.tpm <- t(t(cage_mat.tpm) * sf)

summary(colSums(cage_mat.tpm))
```

```{r}
eRNA.expr.mat <- read.table("~/Resources/eRNA_Fantom5/F5.mm10.enhancers.expression.matrix", header = T, sep = "\t", stringsAsFactors = F)

sum(colnames(cage_mat) == colnames(eRNA.expr.mat))

eRNA.mat.tpm <- t(t(eRNA.expr.mat) / colSums(cage_mat)) * 1e6
eRNA.mat.tpm <- t(t(eRNA.mat.tpm) * sf)

summary(colSums(eRNA.mat.tpm))
```

Cerebellum samples to use for the correlations

```{r}
cereb_samples <- eRNA.stats$sample[grepl("cerebellum, embryo|cerebellum, neonate", eRNA.stats$description)]

eRNA.mat.tpm.cerebellum <- eRNA.mat.tpm[eRNA.gr$eRNA[subjectHits(eRNA_overlaps)], cereb_samples]

cage_mat.tpm.cerebellum <- cage_mat.tpm[cage.gr$cage[subjectHits(cage_prom_overlaps)], cereb_samples]

dim(eRNA.mat.tpm.cerebellum)
dim(cage_mat.tpm.cerebellum)
```

```{r}
gene2peak.correlations <- readRDS("~/Mouse_Cereb/Revisions/002_cre2gene_assignment/CRE_to_gene_assignment_all_tests.rds")

## Filter for pairs matching eRNAs and CAGE TSS
eRNA.peaks <- peak.gr$peak[queryHits(eRNA_overlaps)]
cage.prom <- prom.gr$peak[queryHits(cage_prom_overlaps)]

gene2peak.correlations.inFantom <- filter(gene2peak.correlations, peak %in% eRNA.peaks, prom %in% cage.prom)

nrow(gene2peak.correlations.inFantom)

gene2peak.correlations.inFantom <- left_join(gene2peak.correlations.inFantom, 
                                             data.frame(peak=peak.gr$peak[queryHits(eRNA_overlaps)],
                                                        eRNA=eRNA.gr$eRNA[subjectHits(eRNA_overlaps)],
                                                        stringsAsFactors = F)) %>%
  left_join(data.frame(prom=prom.gr$peak[queryHits(cage_prom_overlaps)],
                                                        cage=cage.gr$cage[subjectHits(cage_prom_overlaps)],
                                                        stringsAsFactors = F))

head(gene2peak.correlations.inFantom)

## We should also filter out eRNAs or cage peaks with 0 activity in the cerebellum matrix
gene2peak.correlations.inFantom <- filter(gene2peak.correlations.inFantom,
                                          eRNA %in% row.names(eRNA.mat.tpm.cerebellum)[rowSums(eRNA.mat.tpm.cerebellum > 0) >= 2],
                                          cage %in% row.names(cage_mat.tpm.cerebellum)[rowSums(cage_mat.tpm.cerebellum > 0) >= 2])

nrow(gene2peak.correlations.inFantom)
```

```{r, fig.width=3, fig.height=4}
gene2peak.correlations.inFantom$fantomCorr <- sapply(1:nrow(gene2peak.correlations.inFantom), function(i) {
  cor(eRNA.mat.tpm.cerebellum[gene2peak.correlations.inFantom$eRNA[i],], 
      cage_mat.tpm.cerebellum[gene2peak.correlations.inFantom$cage[i],])
})

summary(gene2peak.correlations.inFantom$fantomCorr)

p <- ggplot(gene2peak.correlations.inFantom, aes(x=sig.interaction, y=fantomCorr, fill=sig.interaction)) +
  geom_boxplot(notch = T) +
  ylab("eRNA-TSS correlation in bulk\ndeveloping cerebellum (FANTOM5)") +
  scale_fill_manual(values = c("indianred", "deepskyblue3"), guide=F) +
  xlab("Significant interaction") +
  theme_classic()

p

pdf("~/Mouse_Cereb/Revisions/Figures/002_CRE2gene_assignment_FANTOM5_correlations.pdf", width = 2, height = 4, useDingbats = F); print(p); dev.off()
```

```{r}
sessionInfo()
```

